CCDC113 stabilizes sperm axoneme and head-tail coupling apparatus to ensure male fertility

Version 1

Abstract

Abstract

The structural integrity of the sperm is crucial for male fertility, defects in sperm head-tail linkage and flagellar axoneme are associated with acephalic spermatozoa syndrome (ASS) and the multiple morphological abnormalities of the sperm flagella (MMAF). Notably, impaired head-tail coupling apparatus (HTCA) often accompanies defects in the flagellum structure, however, the molecular mechanisms underlying this phenomenon remain elusive. Here, we identified an evolutionarily conserved coiled-coil domain-containing (CCDC) protein, CCDC113, and found the disruption of CCDC113 produced spermatozoa with disorganized sperm flagella and HTCA, which caused male infertility. Further analysis revealed that CCDC113 could bind to CFAP57 and CFAP91, and function as an adaptor protein for the connection of radial spokes, nexin-dynein regulatory complex (N-DRC) and doublet microtubules (DMTs) in sperm axoneme. Moreover, CCDC113 was identified as a structural component of HTCA, collaborating with SUN5 and CENTELIN to connect sperm head to tail during spermiogenesis. Together, our studies reveal that CCDC113 serve as critical hub for sperm axoneme and HTCA stabilization, providing insights into the potential pathogenesis of infertility associated with human CCDC113 mutations.

Introduction

The integrity of spermatozoa is necessary for its migration through the female reproductive tract and successful fertilization, and an intact spermatozoon contains proper sperm head, head-tail coupling apparatus (HTCA), and flagellum (Parker, 2020; Roldan, 2019). Many components have been identified in sperm HTCA or flagellum, and are essential for the sperm integrity (Lehti & Sironen, 2017; Wu et al., 2020). Defects in sperm flagellum and HTCA may lead to reduced sperm motility or aberrant sperm morphology, termed as multiple morphological abnormalities of the sperm flagella (MMAF) or acephalic spermatozoa syndrome (ASS), in turn, which causes male infertility (Sudhakar et al., 2021; Tu et al., 2020).

Sperm flagellum possesses an evolutionarily conserved axonemal structure composed of “9+2” microtubules, specifically, nine peripheral doublet microtubules (DMTs) surrounding two central microtubules known as the central pair (CP) (Inaba & Mizuno, 2016). Axonemal dyneins, radial spokes (RS), and the nexin-dynein regulatory complex (N-DRC), are arranged on DMTs with a 96-nm repeating unit structures (Kumar & Singh, 2021). In axoneme, the N- DRC and RS are required for axonemal integrity, which form a crossbridge between neighboring DMTs and link the DMTs to the central apparatus, respectively (Canty et al., 2021; Ishikawa, 2017; Kumar & Singh, 2021). Recent advances in artificial intelligence, biochemical techniques, and cryo-electron microscopy (cryo-EM) facilitated the analysis of the axonemal structures, and numerous components were detected among RS, N-DRC and DMTs, which may serve as hubs to stabilize the axoneme (Bazan et al., 2021; Leung et al., 2023; Walton et al., 2023; Zhou et al., 2023). Such as, CFAP91 is identified to extend from the base of RS2 through the N-DRC base plate to the RS3, and stabilizes RS2 and RS3 on the DMTs (Bicka et al., 2022; Dymek et al., 2011; Gui et al., 2021). CFAP57 also extends through the N-DRC and interacts with RS3 via its C-terminal region (Ghanaeian et al., 2023). Recent analysis reveals that CCDC96 and CCDC113 could form a complex and extend parallel to the N-DRC and connect the base of RS3 to the tail of dynein g (IDA g) and the N- DRC (Bazan et al., 2021; Ghanaeian et al., 2023). However, the functions of these proteins in sperm flagellum stabilization remain unknown.

Sperm flagellum is tightly anchored to the sperm head through the HTCA, a complex structure based on the centrosome (Wu et al., 2020). This structure consists of two cylindrical microtubule-based centrioles and associated components, including the well-organized segmented columns, along with the capitulum plate and basal plate. The segmented columns and capitulum plate, located below the basal plate, are thought to originate from the dense material emanating from the proximal centriole (Fawcett & Phillips, 1969; Zamboni & Stefanini, 1971). Many proteins have been identified in sperm HTCA, and their mouse models display acephalic spermatozoa syndrome phenotype (Wu et al., 2020). SPATA6 is the first protein identified as a component of the HTCA using the knockout mouse model, and is crucial for the formation of the sperm HTCA (Yuan et al., 2015). Deficiencies in SUN5 (Elkhatib et al., 2017; Fang et al., 2018; Liu et al., 2020; Sha et al., 2018; Shang et al., 2018; Shang et al., 2017; Xiang et al., 2022; Zhang et al., 2021; Zhu et al., 2016) and PMFBP1 (Deng et al., 2022; Liu et al., 2020; Liu et al., 2021; Lu et al., 2021; Nie et al., 2022; Sha et al., 2019; Zhu et al., 2018) have been demonstrated to be associated with ASS in both humans and mice. The centriole-related protein, CENTLEIN, serves as a bona fide linker between SUN5 and PMFBP1 to participates in the HTCA assembly (Zhang et al., 2021). Notably, impaired HTCA often accompanies defects in the sperm flagellum (Hall et al., 2013; Shang et al., 2017; Yuan et al., 2015; Zhang et al., 2021; Zhu et al., 2018), suggesting that the HTCA stabilization may be closely associated with the integrity of sperm flagellum. However, the mechanism that maintains both sperm flagellum and HTCA stabilization remains to be clarified.

Here, we identified an evolutionarily conserved coiled-coil domain-containing (CCDC) protein, CCDC113, and found it could complex with CFAP57 and CFAP91 for the connection of RS, N-DRC and DMTs in axoneme. The knockout of Ccdc113 produced spermatozoa with flagella defects and head-tail linkage detachment, in turn, which caused male infertility. Ultra-structural analyses showed that the depletion of CCDC113 resulted in the destroyed sperm axoneme and HTCA. CCDC113 is localized on the manchette, HTCA and flagellum in elongating and elongated spermatids. Further analysis revealed that CCDC113 is indispensable for the connection of CFAP91 and DRC2 to the DMTs in the sperm axoneme, and CCDC113 interacts with SUN5 and CENTLEIN to stabilize sperm HTCA. All these results suggest that CCDC113 server as a critical hub to maintain the structural integrity of both sperm flagellum and HTCA.

Results

CCDC113 complexes with CFAP57 and CFAP91 for the connection of RS, N-DRC and DMTs

CCDC113 is an evolutionarily conserved coiled-coil domain-containing (CCDC) protein in the ciliated species, and a comparative analysis of the structures of CCDC113 from Tetrahymena thermophila to Homo sapiens showed that CCDC113 orthologs shared structural similarity with each other (Fig. 1A). Recent cryo-EM analysis in the structure of the 96-nm modular repeats of axonemes from the Tetrahymena thermophila cilia and human respiratory cilia revealed that CCDC113 localized the linker region among RS, N-DRC and DMTs (Fig. 1B), indicating it may serve as a structural component connecting RS, N-DRC and DMTs (Bazan et al., 2021; Ghanaeian et al., 2023). To investigate further, we examined the interactions between CCDC113 and its neighboring axoneme-associated proteins, CFAP57 and CFAP91 (Fig. 1B). We transfected HEK293T cells with a GFP-tagged CCDC113 and FLAG-tagged CFAP57 or CFAP91, and then performed anti-FLAG- immunoprecipitations (Fig. 1A). CCDC113 were present in both FLAG-CFAP57 and FLAG- CFAP91 immunoprecipitates (Fig. 1C, D), indicating CCDC113 interacts with both CFAP57 and CFAP91. Given that CFAP91 have been reported to stabilize RS on the DMTs (Bicka et al., 2022; Dymek et al., 2011; Gui et al., 2021) and the cryo-EM analysis shows CCDC113 is closed to DMTs, we speculate CCDC113 may connect RS to DMTs via binding CFAP91 and microtubules. To test it, we detected the interaction between CCDC113 and β-tubulin (TUBB5), and found CCDC113 was present in MYC-TUBB5 immunoprecipitate (Fig. 1E). As CFAP57 extends through the N-DRC and CCDC113 is closed to the N-DRC (Ghanaeian et al., 2023), we further examined the interaction between CCDC113 and N-DRC components that closed to DMTs. The co-immunoprecipitation (co-IP) analysis showed that CCDC113 could bind to DRC1, DRC2, and DRC3 (Fig. 1E-H). Therefore, CCDC113 may be serve as an adaptor protein for the connection of RS, N-DRC and DMTs, and this protein may be a critical hub for axoneme stabilization.

CCDC113 is an evolutionarily conserved axoneme-associated protein.

(A) Multiple species phylogenetic tree of CCDC113. Structural similarity scores (Z scores) of CCDC113 orthologs in Homo sapiens, Mus musculus, Phascolarctos cinereus, Danio rerio, Chlamydomonas reinhardtii and Tetrahymena thermophila were derived through the DALI webserver for pairwise structure comparisons (Holm & Laakso, 2016). (B) Positioning of CCDC113 within the 96-nm repeat of human axoneme (Walton et al., 2023). CCDC113 form a complex with CCDC96, is located at the base of RS3 and adjacent to CFAP91 and CFAP57. CFAP91 originates at the base of RS2 and links the RS3 subunits (CFAP251 and CFAP61). (C-H) Neighboring axoneme-associated proteins were expressed or co-expressed with CCDC113 in HEK293T cells, and the interactions of CCDC113 with CFAP57, CFAP91, TUBB5, DRC1, DRC2 and DRC3 were examined through co-immunoprecipitation. IB, immunoblotting; IP, immunoprecipitation.

CCDC113 is required for male fertility

To investigate the physiological functions of CCDC113, we generated a Ccdc113 knockout mouse strain using the CRISPR/Cas9 system (Fig. 2A). Ccdc113−/−mice were genotyped by genomic DNA sequencing and further confirmed by polymerase chain reaction (PCR). Genotypes were distinguished by a 539 bp band for Ccdc113+/+ mice, a 461 bp band for Ccdc113−/−mice, and two bands of 539 bp and 461 bp for Ccdc113+/– mice (Fig. 2B). Further immunoblotting analysis validated the successful elimination of CCDC113 in total protein extracts from Ccdc113−/− testes (Fig. 2C). Mice lacking Ccdc113 showed no gross abnormalities in appearance or behavior, and no obvious differences in body weight (Fig. 2H). Symptoms such as hydrocephalus or left-right asymmetry defects were also not found in Ccdc113−/−mice (Fig. S1A). Additionally, the deficiency of CCDC113 did not affect ciliogenesis in the lung and trachea (Fig. S1B-D). Subsequently, we assessed the fertility of 2- month-old male and female Ccdc113−/− mice. Ccdc113−/−female mice were able to generate offspring after mating with wild-type (WT) adult males, which was similar to Ccdc113+/+ female mice (Fig. 2D). However, Ccdc113−/− male mice exhibited normal mating behavior with copulatory plugs but failed to produce any offspring after mating with WT adult female mice (Fig. 2E). Thus, the knockout of Ccdc113 results in male infertility.

Ccdc113 knockout leads to male infertility.

(A) The CRISPR-Cas9 strategy for generating the Ccdc113 knockout mice. (B) Genotyping to identify Ccdc113 knockout mice. (C) Immunoblotting of CCDC113 in Ccdc113+/+and Ccdc113−/− testes. TUBULIN served as the loading control. (D) The average litter size of Ccdc113+/+ and Ccdc113−/− female mice in 2 months (n = 5 independent experiments). Data are presented as mean ± SD. ns: indicates no difference. (E) The average litter size of Ccdc113+/+ and Ccdc113−/− male mice in 2 months (n = 5 independent experiments). Data are presented as mean ± SD. ****P < 0.0001. (F) The size of testes was similar in Ccdc113+/+ and Ccdc113−/− mice. (G) The testis weights of Ccdc113+/+ and Ccdc113−/− male mice (n = 5 independent experiments). Data are presented as mean ± SD. ns: indicates no difference. (H) The body weights of Ccdc113+/+and Ccdc113−/− male mice (n = 5 independent experiments). Data are presented as mean ± SD. ns: indicates no difference. (I) The ratio of testis weight/body weight in Ccdc113+/+and Ccdc113−/− male mice (n = 5 independent experiments). Data are presented as mean ± SD. ns: indicates no difference. (J) H&E staining of testes sections from Ccdc113+/+ and Ccdc113−/− male mice. Red arrowheads indicate the abnormal sperm flagellum in the Ccdc113−/− testis seminiferous tubule. (K) Immunofluorescence of acetylated-tubulin (red) in testes sections from Ccdc113−/− male mice show flagellar defects.

Ccdc113 knockout mice produce spermatozoa with flagella defects and head-tail linkage detachment

To further investigate the cause of male infertility, we initially examined Ccdc113-/- testis at both gross and histological levels. Ccdc113 knockout did not affect either testis size (Fig. 2F, G) or the ratio of testis weight to body weight (Fig. 2H, I). Histological sections stained with hematoxylin-eosin (H&E) revealed that seminiferous tubules of Ccdc113+/+ mice exhibited a tubular lumen with flagella emerging from the developing spermatids. In contrast, the flagellum staining appeared reduced in Ccdc113−/− seminiferous tubules (Fig. 2J, red asterisk). Immunofluorescence staining for acetylated tubulin (ac-tubulin), a marker for sperm flagellum (Martinez et al., 2020), further confirmed the defects of flagellum in Ccdc113−/− mice (Fig. 2K). Next, we examined spermatids at different stages in Ccdc113−/− testes by using Periodic acid–Schiff (PAS) staining, and found that the acrosome biogenesis and nucleus morphology in Ccdc113−/− spermatids from step 1 to step 10 were normal compared to Ccdc113+/+ spermatids. However, abnormal club-shaped heads were observed in Ccdc113−/− spermatids from step 11 to step 16 (Fig. S2A, black asterisk). In addition, the manchette of Ccdc113−/− spermatids were more elongated compared to that of Ccdc113+/+ spermatids (Fig. S2B). Therefore, the disruption of CCDC113 impaired spermiogenesis.

Next, we examined the spermatozoa in the caudal epididymis, and found that the sperm count in the Ccdc113−/− caudal epididymis was significantly decreased compared with the control group (Fig. 3A, B). The motility of the released spermatozoa from Ccdc113+/+ and Ccdc113−/− cauda epididymis showed that the Ccdc113−/−spermatozoa were completely immobility (Fig. 3C). H&E staining of the caudal epididymis showed fewer hematoxylin-stained sperm heads in the Ccdc113−/−cauda epididymis compared to that of the Ccdc113+/+ cauda epididymis. Notably, in contrast to the control group, where the epididymal lumen exhibited linear eosin staining, the Ccdc113−/− mouse showed a dramatic presence of coiled eosin-stained structures without sperm heads in the epididymal lumen (Fig. 3A, red circles). To determine the morphological characteristics of the spermatozoa, we conducted single-sperm immunofluorescence using an anti-α/β-tubulin antibody to label the sperm flagellum and lectin peanut agglutinin (PNA) to visualize the sperm acrosome (Nakata et al., 2015). We noticed that Ccdc113−/−spermatozoa showed severe morphological malformations, including sperm head-tail detachment (type 1), abnormal sperm head with curly tail (type 2), normal sperm head with curly tail (type 3) (Fig. 3D, E). Overall, these findings clearly suggest that the deletion of Ccdc113 results in both sperm flagellum deformity and detachment of sperm head-to-tail linkage, and causes a special type of acephalic spermatozoa, which may be responsible for the Ccdc113−/−male infertility.

Ccdc113 knockout results in sperm flagella defects and sperm head-tail detachment. (A) H&E staining of the caudal epididymis from Ccdc113+/+and Ccdc113−/− male mice in 2 months. Red circles indicate coiled eosin-stained structures without sperm heads in the epididymal lumen. (B) Analysis of sperm counts in Ccdc113+/+ and Ccdc113−/−male mice (n = 5 independent experiments). Mature spermatozoa were extracted from unilateral cauda epididymis and dispersed in PBS. Sperm counts were measured by hemocytometers. Data are presented as the mean ± SD. ****P < 0.0001. (C) Motile sperm in Ccdc113+/+ and Ccdc113-/-mice (n = 5 independent experiments). Data are presented as mean ± SD. ****P < 0.0001. (D) Ccdc113+/+ and Ccdc113−/− spermatozoa were co-stained with a flagellar marker α/β-tubulin (red) and an acrosomal marker PNA. Nuclei were stained with DAPI (blue). (E) Quantification of different categories of Ccdc113+/+, Ccdc113−/− spermatozoa (n = 3 independent experiments). Data are presented as the mean ± SD.

CCDC113 localizes on the sperm neck and flagellum regions

To gain further insights into the functional role of CCDC113 during spermiogenesis, we detected the expression of CCDC113, and found that CCDC113 was predominantly expressed in mouse testis (Fig. 4A). Furthermore, CCDC113 was first detected in testis at postnatal day 7 (P7), and the level of CCDC113 increased continuously from P21 onward, with the highest levels detected in adult testes (Fig. 4B), indicating CCDC113 might be highly expressed during spermiogenesis. Then, we conducted immunofluorescence analysis of CCDC113 in Ccdc113+/+ and Ccdc113−/− germ cells to characterize its precise localization during spermatogenesis (Fig. S3). CCDC113 showed punctum single near the nuclei of spermatocyte and round spermatids, and localized on the manchette, sperm neck and flagellum regions in elongating and elongated spermatids (Fig. S3).

CCDC113 localizes to the HTCA, manchette and sperm flagellum.

(A) CCDC113 was predominately expressed in testis and slightly in lung. Immunoblotting of CCDC113 was performed in the spleen, intestine, lung, thymus, testis and ovary. Asterisks indicate unspecific bands. TUBULIN served as the loading control. (B) CCDC113 was expressed starting in P7 testes. TUBULIN served as the loading control. Asterisks indicate unspecific bands. (C) Immunofluorescence of CCDC113 (red) and CENTRIN1/2 (green) in developing germ cells. CCDC113 partially colocalize with centriolar protein CENTRIN1/2. (D) Immunofluorescence of CCDC113 (red) and α/β-tubulin (green) in developing germ cells. The manchette was stained with the anti-α/β-tubulin antibody. (E-F) CCDC113 localizes to the HTCA and flagellum in mature mouse spermatozoa and human spermatozoa. Nuclei were stained with DAPI (blue).

To further confirm it, we co-stained CCDC113 and α/β-tubulin, which represented the manchette and flagellum in spermatids (Lehti & Sironen, 2016) (Fig. 4D). The immunofluorescence analysis showed that CCDC113 localized on the manchette surrounding the spermatid head from step 9 to step 14, as well as the testicular sperm neck and flagellum (Fig. 4D). Given that CCDC113 was initially identified as a component of centriolar satellites (Firat-Karalar et al., 2014), the punctum single of CCDC113 in spermatocyte and spermatids may localize around the centrosome. To test it, we performed the immunofluorescent staining of CCDC113 and centrosomal protein CENTRIN1/2 in spermatocytes and spermatids, and found the signal of CCDC113 could partially colocalize with CENTRIN1/2 (Fig. 4C). Thus, CCDC113 localizes on the centrosome, manchette, sperm neck and flagellum regions in the developing germ cells.

Next, we examined the localization of CCDC113 in mature spermatozoa, and found that CCDC113 was localized in the sperm neck and flagellum regions (Fig. 4E). Similar localization of CCDC113 could also be detected in human mature spermatozoa (Fig. 4F).

Therefore, the localization of CCDC113 on the sperm neck and flagellum is persistent in mature spermatozoa, which may be essential for the integrity of sperm flagellum and head to tail.

Ccdc113 knockout results in the disorganization of the sperm flagellum structures

To delineate the sperm flagella defects in Ccdc113−/−mice, we conducted transmission electron microscopy (TEM) examination of longitudinal sections of Ccdc113−/− spermatozoa. The TEM analysis revealed a significant presence of unremoved cytoplasm, including disrupted mitochondria, damaged axonemes, and large vacuoles in Ccdc113−/−spermatozoa (Fig. 5A, red asterisks). Cross sections of the principal piece of Ccdc113−/−spermatozoa further revealed partial loss or unidentifiable “9+2” axonemal structures, along with the disruption of the fibrous sheath and outer dense fibers (Fig. 5A). We further examined the axonemal structure in Ccdc113−/− testicular spermatids using TEM, and disorganized axonemal microtubules were detected in Ccdc113−/−testicular spermatids (Fig. 5B). In contrast to the regular positioning of the CP and nine peripheral DMTs in the Ccdc113+/+spermatid axoneme, the Ccdc113−/− spermatids exhibited a scattered arrangement of DMTs, and no distinct radial spokes were observed (Fig. 5B, red arrowheads). Diffused axonemal signals could also be observed in Ccdc113−/− testicular germ cells by using ac-tubulin immunofluorescence staining during spermiogenesis (Fig. 5B, Fig. S2C). These results indicated CCDC113 is essential for the integrity of sperm flagellum.

CCDC113 is indispensable for the docking of CFAP91 and DRC2 to the DMTs to maintain the structural integrity of the axoneme. (A) Transmission electron microscopy (TEM) analysis of spermatozoa from the cauda epididymidis of Ccdc113+/+and Ccdc113−/− male mice. The flagellar longitudinal sections of Ccdc113−/− spermatozoa revealed unremoved cytoplasm, including disrupted mitochondria, damaged axonemes. and large vacuoles. Asterisks indicate large vacuoles. Cross sections of the principal piece of Ccdc113−/− spermatozoa further revealed partial loss or unidentifiable “9+2” structures, along with the disruption of the fibrous sheath and outer dense fibers. (B) Transmission electron microscopy (TEM) analysis of the axoneme in testicular spermatids from Ccdc113+/+ and Ccdc113−/− male mice. The red arrowheads indicate the absence of significant radial spokes (RSs). MS: mitochondrial sheath, Mi: mitochondrial, AX: axoneme. FS: fibrous sheath. DMT: doublet microtubule, MT: microtubule, CP: central pair, ODF: outer dense fiber, RS: radial spokes. (C) The immunofluorescence analysis for CFAP91 (green) and α/β-tubulin (red) was performed in Ccdc113+/+and Ccdc113−/− spermatozoa. Nuclei were stained with DAPI (blue). White asterisks indicate regions that are not co-located with tubulin. (D) The immunofluorescence analysis for DRC2 (green) and α/β-tubulin (red) was performed in Ccdc113+/+ and Ccdc113−/− spermatozoa. Nuclei were stained with DAPI (blue). White asterisks indicate regions that are not colocated with tubulin. (E-F) Line-scan analysis (white line) using the Image J software.

CCDC113 has been shown to localize at the base of the RS3 and interact with neighboring axoneme-associated proteins (Fig. 1B-H). Since obvious disorganized “9+2” axonemal structure was detected in the cross-sectioned Ccdc113−/−flagellar specimen (Fig. 5A, B), we speculated that CCDC113 likely served as an adaptor to connect the neighboring axoneme- associated proteins to DMTs. Pursuing this, we examined the flagellar localization of CFAP91 in Ccdc113−/− spermatozoa, which is positioned in close proximity to CCDC113 at the root region of RS3, critical for the localizations of calmodulin-associated and spoke-associated complex (CSC) proteins CFAP61 and CFAP251 (Bicka et al., 2022; Meng et al., 2024).

Immunofluorescence results indicated that the absence of Ccdc113 leads to the abnormal distribution of CFAP91 on the axoneme, where CFAP91 could not colocalize with DMTs. (Fig. 5C, white asterisks and 5E). Given that DRC2 serves as the core component of the axonemal N-DRC (Jreijiri et al., 2023) and CCDC113 could bind to DRC2 (Fig 1G), we further detected DRC2 localization in the Ccdc113+/+ and Ccdc113−/−spermatozoa. Immunofluorescence analysis showed that DRC2 exhibited distinct signals that were not colocalized with DMTs of Ccdc113−/−spermatozoa (Fig. 5D, white asterisks and 5F). These findings collectively indicate that CCDC113 is indispensable for the connection of CFAP91 and DRC2 to the DMTs, which is required for structural integrity of the sperm axoneme.

Ccdc113 knockout impairs head-to-tail anchorage of the spermatids

To address how Ccdc113 knockout caused acephalic spermatozoa, we initially examined wherein the flagellum detached from sperm head in Ccdc113−/− mice. The proportion of decapitated tails in the Ccdc113−/− caput, corpus and caudal of the epididymis exhibited similar proportions (Fig. 6A), indicating the separation of the sperm head and tail in Ccdc113−/−mice may occur either within the seminiferous tubules or upon entering the caput of the epididymis. To confirm it, we performed PAS staining to examine the stages of spermiogenesis in Ccdc113+/+ and Ccdc113−/−testes (Fig. 6B). We found that Ccdc113+/+ sperm heads in stages VII-VIII were oriented towards the basal membrane, whereas Ccdc113−/− sperm heads were oriented towards the tubule lumen during stages VII-VIII (Fig. 6C, arrows), which may be attributed to the separation of sperm heads from flagellum during spermiogenesis. Additionally, mature sperm heads were still observable at stages IX–X in Ccdc113−/− testes, while mature spermatozoa were released into the lumen of the seminiferous tubule at stage VIII in Ccdc113+/+testes (Fig. 6B, red asterisk). These results suggest that the sperm head and flagellum might undergo separation during spermiation in the Ccdc113−/− testes.

Ccdc113 knockout spermatids display impaired HTCA. (A) The proportion of decapitated tails in Ccdc113+/+ and Ccdc113−/− corpus, caput and cauda epididymis (n=3 independent experiments). Data are presented as the mean ± SD. ****P < 0.0001. ***P < 0.001. (B) PAS staining of testes sections from Ccdc113+/+ and Ccdc113−/− mice. The green arrows indicate the orientation of the sperm heads in stages V–VIII seminiferous epithelia. Ccdc113−/− sperm head could still be detected in stages IX–X seminiferous epithelia. P: pachytene spermatocyte, spz: spermatozoa, rSt: round spermatid, eSt: elongating spermatid, Z: zygotene spermatocyte, M: meiotic spermatocyte. (C) Ccdc113−/− spermatids lost their head orientation toward the basement membrane during spermiation in stages VII–VIII seminiferous epithelia. L: lumen, B: basement membrane. (D) Defective HTCA formation in Ccdc113−/−spermatids. TEM analyses of the stepwise development of the HTCA were performed in Ccdc113+/+ and Ccdc113−/− testes. In Ccdc113+/+ spermatids, the well-defined coupling apparatus was tightly attached to the sperm head. In Ccdc113−/− spermatids, segmented columns (Scs), the capitulum (Cp) were absent. The red asterisks indicate the distance between the sperm head and HTCA. The white arrows indicate the dense material surrounding the proximal centriole. Nu: nuclear, Bp: basal plate, Cp: capitulum, Sc: segmented column, Pc: proximal centriole, Dc: distal centriole, An: annulus, Ax: axoneme, Rn: redundant nuclear envelope.

Next, we examined the development of the HTCA in Ccdc113+/+and Ccdc113−/− spermatids using TEM. In Ccdc113+/+ step 9–11 spermatids, the well-defined coupling apparatus consisted of basal plate, capitulum plate, segmented columns, proximal centriole, distal centriole was tightly attached to the sperm head. However, in Ccdc113−/− step 9–11 spermatids, the abnormal HTCA detached from the sperm head (Fig.6D, red asterisk). Further observation of the HTCA structure revealed the absence of segmented columns and capitulum plate; only dense material surrounding the proximal centriole and basal plate could be detected (Fig. 6D, white arrow). The basal plates were abnormally distant from their native implantation site on the nucleus of Ccdc113−/− elongating and elongated spermatids (Fig. 6D). Taken together, our results indicate that the disruption of Ccdc113 causes the destroyed coupling apparatus detachment from the sperm head during spermiogenesis, and CCDC113 is required for the integrity of the sperm HTCA.

CCDC113 cooperates with SUN5 and CENTLEIN to stabilize sperm HTCA

To investigate the molecular function of CCDC113 in sperm head-tail linkage, we examined interaction between CCDC113 and reported HTCA-associated proteins, including SUN5, CENTLEIN, PMFBP1, and SPATA6 (Shang et al., 2017; Yuan et al., 2015; Zhang et al., 2021; Zhu et al., 2018). GFP-tagged CCDC113 and FLAG-tagged HTCA-associated proteins were co-transfected into HEK293T cells, followed by the immunoprecipitation with anti-GFP antibody. We found that CCDC113 could interact with SUN5 and CENTLEIN, but not with PMFBP1 and SPATA6 (Fig. 7A-C, F). We further conducted co-immunoprecipitation assays in the reverse direction and verified that both SUN5 and CENTLEIN could bind to CCDC113 (Fig. 7D, E). To further investigate the localization of CCDC113 in the HTCA, we co-stained mature spermatozoa with antibodies against CCDC113 and SUN5, which has been reported to be localized on the root connecting the HTCA to the nuclear envelope (Shang et al., 2017; Zhang et al., 2021). CCDC113 was located below SUN5, showed partial overlap with SUN5 (Fig. 7G). The centriolar protein CENTLEIN was detected to localize on the HTCA and acted as the critical linker protein between SUN5 and PMFBP1 in the elongating and elongated spermatid (Zhang et al., 2021)). Given that CENTLEIN disappears in mature spermatozoa (Zhang et al., 2021), we performed the immunofluorescent staining of CCDC113 and CENTLEIN in elongated spermatids, and found that CCDC113 could partially colocalize with CENTLEIN at the HTCA (Fig. 7H). Thus, CCDC113 could interact with both SUN5 and CENTLEIN, and localize on the sperm HTCA.

CCDC113 interact with SUN5 and CENTLEIN, participating in sperm head-tail linkage.

(A-C, F) HTCA-associated proteins (SUN5, CENTLEIN, PMFBP1, SPATA6) were expressed or co-expressed with CCDC113 in HEK293T cells, and the interactions of CCDC113 with SUN5, CENTLEIN were examined through co-immunoprecipitation. CCDC113 did not interact with PMFBP1 and SPATA6. IB, immunoblotting; IP, immunoprecipitation. (D) SUN5 interacted CCDC113. pCDNA-FLAG-Ccdc113 and pEGFP- GFP-Sun5 were transfected into HEK293T cells. At 48 h after transfection, the cells were collected for immunoprecipitation (IP) with anti-GFP antibody and analyzed with anti-FLAG and anti-GFP antibodies. (E) CENTLEIN interacted CCDC113. pCMV-FLAG-Centlein and pEGFP-GFP-Sun5 were transfected into HEK293T cells. At 48 h after transfection, the cells were collected for immunoprecipitation (IP) with anti-FLAG antibody and analyzed with anti- FLAG and anti-GFP antibodies. (G) Immunofluorescence of CCDC113 (red) and SUN5 (green) in mature spermatozoa. Nuclei were stained with DAPI (blue). (H) Immunofluorescence of CCDC113 (red) and CENTLEIN (green) in testicular step13-step14 spermatid. Nuclei were stained with DAPI (blue). (I) Immunofluorescence analysis for SPATA6 (green) and α/β-tubulin (red) was performed in Ccdc113+/+ and Ccdc113−/− spermatozoa. Nuclei were stained with DAPI (blue). (J) Quantification ratio of SPATA6 on the detached sperm tail (n =3 independent experiments). At least 200 spermatozoa were analyzed for each mouse. (K) Quantification ratio of CCDC113 on the detached sperm tail (n =3 independent experiments). At least 200 spermatozoa were analyzed for each mouse. (L) Immunofluorescence analysis for CCDC113 (red) was performed in wild type (WT), Sun5−/− and Centlein−/− spermatozoa. Nuclei were stained with DAPI (blue).

The HTCA localization of CCDC113 may be response for the integrity of HTCA. To test it, we detected the localization of classical HTCA component, SPATA6, in Ccdc113−/−and Ccdc113+/+spermatozoa by performing the immunofluorescent staining. We found that SPATA6 was not attached to the implantation fossa of the sperm nucleus (Fig. 7I, J), indicating that CCDC113 is essential for the integrity of sperm HTCA. To further understand the functional relationships among CCDC113, SUN5, and CENTLEIN at the sperm HTCA, we examined the localization of CCDC113 in Sun5-/- and Centlein−/− spermatozoa. Compared to the control group, CCDC113 predominantly localized on the decapitated flagellum in both Sun5-/-and Centlein−/− spermatozoa (Fig. 7K, L), indicating SUN5 and CENTLEIN are indispensable for the docking of CCDC113 to the implantation site on the sperm head. Taken together, these data demonstrate that CCDC113 cooperate with SUN5 and CENTLEIN to stabilize sperm HTCA and anchor the sperm head to the tail.

Discussion

CCDC113 is a highly evolutionarily-conserved component of motile cilia/flagella. Studies in the model organism Tetrahymena thermophila have revealed that CCDC113 connects RS3 to dynein g and the N-DRC, which plays essential role in cilia motility (Bazan et al., 2021; Ghanaeian et al., 2023). However, the physiological functions of CCDC113 in mammals remain unknown. In this study, we reveal that CCDC113 is indispensable for male fertility, as Ccdc113 knockout mice produce spermatozoa with flagella defects and head-tail linkage detachment (Fig. 3D). CCDC113 is localized on the sperm neck and flagellum regions in the elongating and elongated spermatids. In sperm flagellum, CCDC113 could interact with both CFAP57 and CFAP91, and serve as an adaptor protein for the connection of RS, N-DRC and DMTs to stabilize sperm flagellum. In sperm head to tail connecting piece, CCDC113 could bind to SUN5 and CENTLEIN to stabilize sperm HTCA and anchor the sperm head to the tail. Thus, CCDC113 is essential for the integrity of both sperm axoneme and sperm HTCA.

Recent cryo-EM analysis in the axonemes from Tetrahymena thermophila cilia and human respiratory cilia revealed that CCDC113 localized the linker region among RS3, N-DRC and DMTs (Bazan et al., 2021; Ghanaeian et al., 2023), and we found that CCDC113 indeed interacted with its neighboring axoneme-associated proteins, CFAP57 and CFAP91 (Fig. 1C, D). CFAP57 has been identified as the adaptor protein responsible for assembling dynein g and d, and it can interact with both N-DRC and RS3 (Ghanaeian et al., 2023; Ma et al., 2023). Previous studies demonstrated that CFAP91 extended from the base of RS2 through the N- DRC base plate to RS3, playing a crucial role in stabilizing and localizing RS2 and RS3 on the DMT (Bicka et al., 2022; Dymek et al., 2011; Gui et al., 2021). The CFAP91 ortholog, FAP91, interacts with three N-DRC subunits (DRC1, DRC2, and DRC4), participating in the docking of the N-DRC in Chlamydomonas (Gui et al., 2021). In humans, pathogenic mutations in CFAP91 and DRC2 disrupt sperm flagellum structure, result in MMAF (Jreijiri et al., 2023; Martinez et al., 2020). TEM and immunostaining experiments in spermatozoa showed severe CP and radial spokes defects in CFAP91 mutant patients (Martinez et al., 2020). In this study, we found that the absence of CCDC113 results in severe axonemal disorganization characterized by defective RSs, scattered DMTs, and misplaced CP. Further analysis demonstrated that the deficiency of CCDC113 disrupts the localization of CFAP91 and DRC2 on DMTs. Thus, CCDC113 may function as an adaptor protein to stabilize CFAP91 and DRC2 on the DMT, which facilitates the docking of RS and N-DRC to DMTs, thereby maintaining the integrity of sperm axoneme.

Recent analysis has revealed that certain centrosomal proteins play crucial roles in the assembly and maintenance of sperm HTCA (Avasthi et al., 2013; Liska et al., 2009; Zhang et al., 2021). In early round spermatids, the centriole pair initially localizes to the caudal nuclear pole and expands the electron-dense material, part of which shows striation (Wu et al., 2020). Along with the development of spermatids, the dense material around the centrioles gradually transforms into a well-organized structure, clearly identified as basal plate, capitulum plate and segmented columns (Dooher & Bennett, 1973; Wu et al., 2020). CCDC113 was initially identified through the isolation of centriolar proteins from bovine sperm (Firat-Karalar et al., 2014). During spermiogenesis, CCDC113 could colocalize with CENTRIN1/2 at the centrosome in developing spermatids, and continue to localize on the sperm neck region in elongating and elongated spermatids (Fig. 4C, D). In addition, CCDC113 could bind to HTCA associated centrosomal protein, CENTLEIN, and the disruption of Centlein impaired the attachment of CCDC113 to sperm head. In the Ccdc113−/−spermatids, the capitulum plate and segmented columns were absent, and the basal plate were detached from the implantation site on the nucleus of Ccdc113−/− elongating and elongated spermatids during spermiogenesis (Fig. 6D). These observations raise the possibility that CCDC113 might be a HTCA associated centrosomal protein, and play a role in maintaining the structural integrity of the HTCA.

SUN5 is a transmembrane protein located in the nuclear envelope, and acts as the root that connects the HTCA to the sperm nuclear envelope (Shang et al., 2017). CENTLEIN could directly bind to SUN5 and PMFBP1, and works as a linker between SUN5 and PMFBP1 to maintain the integrity of HTCA (Zhang et al., 2021). CCDC113 could interact with SUN5 and CENTLEIN, but not PMFBP1 (Fig. 7A-C), and CCDC113 was in the cytoplasm in Sun5−/− and Centlein−/−spermatozoa (Fig. 7L, K). In addition, CCDC113 colocalizes with SUN5 in the HTCA region, and the immunofluorescence staining in spermatozoa shows that SUN5 is closer to the sperm nucleus than CCDC113 (Fig. 7G, H). Therefore, SUN5 and CENTLEIN may be more closed to the sperm nucleus compared with CCDC113.

Overall, we identified CCDC113 as a structural component of both the flagellar axoneme and the HTCA, playing dual roles in stabilizing the sperm axonemal structure and maintaining the structural integrity of HTCA. As Ccdc113−/−mice did not exhibit other ciliopathies, such as situs inversus, hydrocephalus, and abnormal ciliogenesis of tracheal cilia (Fig. S1A-D), indicating CCDC113 may specially function in spermiogenesis. Given the evolutionary conservation of CCDC113 in humans, we speculate that mutations in CCDC113 may be present in individuals with MMAF or ASS. Men with sperm flagella defects and head-tail linkage detachment may be a new type of ASS.

Materials and Methods

Phylogenetic analysis and structural similarity analysis

Amino acid sequence of CCDC113 of 7 species were downloaded from UniProt. The phylogenetic trees were constructed using MEGA 10.0 (Kumar et al., 2018) with the Neighbor-Joining (NJ) method (Saitou & Nei, 1987). 3D structures of CCDC113 orthologs of 7 species were obtained from AlphaFold Protein Structure Database (Varadi et al., 2022). Structural similarity Z scores were derived through the DALI webserver for all against all structure comparison (Holm & Laakso, 2016).

Animals

The mouse Ccdc113 gene is 44.81 kb and contains 9 exons. Exon 2 to exon 6 of Ccdc113 was chosen as the target site. The knockout mice were generated by CRISPR-Cas9 system from Cyagen Biosciences. The gRNA and Cas9 mRNA were co-injected into fertilized eggs of C57BL/6J mice to generate Ccdc113+/–mice with a 10215 bp base deletion and 64 bp insertion. The resulting heterozygotes were interbred to obtain offspring, which further were genotyped by genomic DNA sequencing to obtain widetype and homozygote mice. The genotyping primers for knockout were: F1: 5’-TCAAATCATCACACCCTGCCTCT-3’, R: 5’-GCTTGCACTCGGGTGATACATAA-3’, and for WT mice, the specific primers were: F2: 5’-CAGGTTCTCAACACCTACAAAGTA-3’, R: 5’-GCTTGCACTCGGGTGATACATAA-3’.

All the animal experiments were performed according to approved institutional animal care and use committee (IACUC) protocols (# 08-133) of the Institute of Zoology, Chinese Academy of Sciences.

Human adult sperm sample preparation

The sperm donation candidates in this study were healthy young Chinese men. Each participant underwent a thorough medical examination and completed an extensive medical and social questionnaire to ensure the exclusion of individuals with genetic or significant medical issues, as outlined in the Basic Standard and Technical Norms of Human Sperm Bank published by the Chinese Ministry of Health. Individuals who smoked, abused drugs, or were heavy drinkers were also excluded from the study. Those who remained eligible signed informed consent forms for voluntary sperm donation and agreed to reside in Beijing for a minimum of six months. The sperm bank documented each participant’s age, date of birth, and date of semen collection. Ethical approval for this study was granted by the Reproductive Study Ethics Committee of Peking University Third Hospital (2017SZ-035). Semen samples were processed using a 40% density gradient of PureSperm (Nidacon International, Molndal, Sweden) through centrifugation at room temperature (500 × g, 30 min) and washed three times with phosphate-buffered saline (PBS). The obtained spermatozoa were utilized for immunofluorescence staining.

Antibodies

Rabbit anti-CCDC113 generated from Dia-an Biotech (Wuhan, China) was diluted at 1:500 for western blotting and a 1: 25 dilution for immunofluorescence. Mouse anti-α-TUBULIN antibody (AC012, Abclonal) was used at a 1:5000 dilution for western blotting. Mouse anti- GFP antibody (M20004, Abmart) was used at a 1:2000 dilution for western blotting. Rabbit anti-MYC antibody (BE2011, EASYBIO) was used at a 1:2000 dilution for western blotting. Rabbit anti-FLAG antibody (20543-1 AP, Proteintech) was used at a 1:2000 dilution for western blotting. Mouse anti-ac-TUBULIN antibody (T7451, Sigma-Aldrich) was used at a 1:200 dilution for immunofluorescence. Mouse anti-α/β-TUBULIN antibody (ab44928, Abcam) was used at a 1: 100 dilution for immunofluorescence. Rabbit anti-DRC2 antibody (NBP2-84617, Novus) was used at a 1:100 dilution for immunofluorescence. Rabbit anti- CFAP91 antibody (bs-9823R, Bioss) was used at a 1:100 dilution for immunofluorescence. In-house-generated mouse anti-SUN5 antibody targeting the SUN5 SUN domain (aa193–373) was used at a 1:100 dilution for immunofluorescence analysis. Rat anti-CENTLEIN antibody was generated by Absea Biotechnology Ltd (Beijing, China) was diluted at a 1:20 dilution for immunofluorescence. Rabbit anti-SPATA6 antibody (11849 -1 AP, Proteintech) was used at a 1:100 dilution for immunofluorescence. The Alexa Fluor 488 conjugate of lectin PNA (1:400, L21409, Thermo Fisher) was used for immunofluorescence. The secondary antibodies were goat anti-rabbit FITC (1:200, ZF-0311, Zhong Shan Jin Qiao), goat anti-rabbit TRITC (1:200, ZF-0316, Zhong Shan Jin Qiao), goat anti-mouse FITC (1:200, ZF-0312, Zhong Shan Jin Qiao) and goat anti-mouse TRITC (1:200, ZF0313, Zhong Shan Jin Qiao).

Sperm motility and sperm count assays

The cauda epididymis was isolated from 8 weeks of mice. Sperm were released in phosphate- buffered saline (PBS, Gibco, C14190500BT) from the incisions of the cauda epididymis for 10 min at 37 ℃. And then the swim-up suspension was used for the analysis of sperm motility with a microscope through a 20x phase objective. Viewing areas in each chamber were imaged using a CCD camera. The samples were analyzed via computer-assisted semen analysis (CASA) using the Minitube Sperm Vision Digital Semen Evaluation System (12500/1300, Minitube Group, Tiefenbach, Germany) and were also analyzed by a computer- aided semen analysis system (CASA). The incubated sperm number was counted with a hemocytometer.

Histology staining

As previously reported (Wang et al., 2018), the testes and caudal epididymis were dissected after euthanasia, and fixed with Bouin’s fixative for 24 h at 4 ℃, then the testes were dehydrated with graded ethanol and embedded in paraffin. For histological analysis, the 5 μm sections were cut and covered on glass slides. Sections were stained with H&E and PAS for histological analysis after deparaffinization.

Electron microscopy analysis

The cauda epididymis and testis were dissected and fixed in 2.5% (vol/vol) glutaraldehyde in 0.1 M cacodylate buffer at 4℃ overnight. After washing in 0.1 M cacodylate buffer, samples were cut into small pieces of approximately 1 mm3, then immersed in 1% OsO4 for 1 h at 4℃. Samples were dehydrated through a graded acetone series (50%, 60%, 70%, 80%, 90%, 95%, 100%) and embedded in resin (DDSA, NMA, enhancer, 812) for staining. Ultrathin sections were cut and stained with uranyl acetate and lead citrate. Images were acquired and analyzed using a JEM-1400 transmission electron microscope.

Scanning electron microscopy

The trachea was fixed in 2.5% glutaraldehyde solution overnight, and dehydrated in graded ethanol, subjected to drying and coated with gold. The images were acquired and analyzed using SU8010 scanning electron microscope.

Immunofluorescence staining

The testis albuginea was peeled and incubated with collagenase IV and hyaluronidase in PBS for 15 min at 37℃, then washed twice with PBS. Next, fixed with 4% PFA for 5 min, and then coated on slide glass to dry out. The slides were washed with PBS three times and then treated with 0.5% TritonX-100 for 5 min, and blocked with 5% BSA for 30 min. Added the primary antibodies and incubated at 4℃ overnight, followed by incubating with a second antibody and DAPI. The images were taken using LSM880 and SP8 microscopes.

The mouse testis was immediately dissected and fixed with 2% paraformaldehyde in 0.05% PBST (PBS with 0.05% Triton X-100) at room temperature for 5 min. The fixed sample was placed on a slide glass and squashed by placing a cover slip on top and pressing down. The sample was immediately flash frozen in liquid nitrogen, and the slides were stored at −80°C for further immunofluorescence experiments. After removing the coverslips, the slides were washed with PBS three times and then treated with 0.1% Triton X-100 for 10 min, rinsed three times in PBS, and blocked with 5% bovine serum albumin (Amresco, AP0027). The primary antibody was added to the sections and incubated at 4°C overnight, followed by incubation with the secondary antibody. The nuclei were stained with DAPI. The immunofluorescence images were taken immediately using an LSM 780 microscope (Zeiss) or SP8 microscope (Leica).

Spermatozoa were released from the cauda epididymis in PBS at 37°C for 15 min, then were spread on glass slides for morphological observation or immunostaining. After air drying, spermatozoa were fixed in 4% PFA for 5 min at room temperature, and slides were washed with PBS three times and blocked with 5% BSA for 30 min at room temperature. The primary antibodies were added to the sections and incubated at 4°C overnight, followed by incubation with the secondary antibody. The nuclei were stained with DAPI and images were taken using an LSM 880 microscope (Zeiss) or SP8 microscope (Leica).

Immunoprecipitation

Transfected cells were lysed in a lysis buffer (50mM HEPES, PH 7.4, 250mM NaCl, 0.1% NP-40 containing PIC and PMSF) on ice for 30 min and centrifugated at 12000 rpm at 4℃ for 15 min, cell lysates were incubated with primary antibody about 12 hours at 4°C and then incubated with protein A-Sepharose (GE, 17-1279-03) for 3 hours at 4°C, then washed three times with lysed buffer and subjected to immunoblotting analysis.

Statistical analysis

All the experiments were repeated at least three times, and the results are presented as the mean ± SD. The statistical significance of the differences between the mean values for the different genotypes was measured by the Student’s t-test with a paired, 2-tailed distribution. The data were considered significant when the P-value was less than 0.05(*), 0.01(**), 0.001(***) or 0.0001(****).

eLife Assessment

This study presents a valuable finding on sperm flagellum and HTCA stabilization. The evidence supporting the authors' claims is incomplete. The work will be of broad interest to cell and reproductive biologists working on cilium and sperm biology.

Reviewer #1 (Public Review):

In this paper, Wu et al. investigated the physiological roles of CCDC113 in sperm flagellum and HTCA stabilization by using CRISPR/Cas knockouts mouse models, co-IP, and single sperm imaging. They find that CCDC113 localizes in the linker region among radial spokes, the nexin-dynein regulatory complex (N-DRC), and doublet microtubules (DMTs) RS, N-DRC, and DMTs and interacts with axoneme-associated proteins CFAP57 and CFAP91, acting as an adaptor protein that facilitates the linkage between RS, N-DRC, and DMTs within the sperm axoneme. They show the disruption of CCDC113 produced spermatozoa with disorganized sperm flagella and CFAP91, DRC2 could not colocalize with DMTs in Ccdc113-/- spermatozoa. Interestingly, the data also indicate that CCDC113 could localize on the HTCA region, and interact with HTCA-associated proteins. The knockout of Ccdc113 could also produce acephalic spermatozoa. By using Sun5 and Centlein knockout mouse models, the authors further find SUN5 and CENTLEIN are indispensable for the docking of CCDC113 to the implantation site on the sperm head. Overall, the experiments were designed properly and performed well to support the authors' observation in each part. Furthermore, the study's findings offer valuable insights into the physiological and developmental roles of CCDC113 in the male germ line, which can provide insight into impaired sperm development and male infertility. The conclusions of this paper are mostly well supported by data, but some points need to be clarified and discussed.

(1) In Figure 1, a sperm flagellum protein, which is far away from CCDC113, should be selected as a negative control to exclude artificial effects in co-IP experiments.

(2) Whether the detachment of sperm head and tail in Ccdc113-/- mice is a secondary effect of the sperm flagellum defects? The author should discuss this point.

(3) Given that some cytoplasm materials could be observed in Ccdc113-/- spermatozoa (Fig. 5A), whether CCDC113 is also essential for cytoplasmic removal?

(4) Although CCDC113 could not bind to PMFBP1, the localization of CCDC113 in Pmfbp1-/- spermatozoa should be also detected to clarify the relationship between CCDC113 and SUN5-CENTLEIN-PMFBP1.

Reviewer #2 (Public Review):

Summary:

In the present study, the authors select the coiled-coil protein CCDC113 and revealed its expression in the stages of spermatogenesis in the testis as well as in the different steps of spermiogenesis with expression also mapped in the different parts of the epididymis. Gene deletion led to male infertility in CRISPR-Cas9 KO mice and PAS staining showed defects mapped in the different stages of the seminiferous cycle and through the different steps of spermiogenesis. EM and IF with several markers of testis germ cells and spermatozoa in the epididymis indicated defects in flagella and head-to-tail coupling for flagella as well as acephaly. The authors' co-IP experiments of expressed CCDC113 in HEK293T cells indicated an association with CFAP91 and DRC2 as well as SUN5 and CENTLEIN.

The authors propose that CCDC113 connects CFAP91 and DRC2 to doublet microtubules of the axoneme and CCDC113's association with SUN5 and CENTLEIN to stabilize the sperm flagellum head-to-tail coupling apparatus. Extensive experiments mapping CCDC13 during postnatal development are reported as well as negative co-IP experiments and studies with SUN5 KO mice as well as CENTLEIN KO mice.

Strengths:

The authors provide compelling observations to indicate the relevance of CCDC113 to flagellum formation with potential protein partners. The data are relevant to sperm flagella formation and its coupling to the sperm head.

Weaknesses:

The authors' observations are consistent with the model proposed but the authors' conclusions for the mechanism may require direct demonstration in sperm flagella. The Walton et al paper shows human CCDC96/113 in cilia of human respiratory epithelia. An application of such methodology to the proteins indicated by Wu et al for the sperm axoneme and head-tail coupling apparatus is eagerly awaited as a follow-up study.

Author response:

The following is the authors’ response to the original reviews.

This study presents a valuable finding on sperm flagellum and HTCA stabilization. The evidence supporting the authors' claims is incomplete. The work will be of broad interest to cell and reproductive biologists working on cilium and sperm biology.

We thank the Editor and the two reviewers for their time and thorough evaluation of our manuscript. We greatly appreciate their valuable guidance on improving our study. In the revised manuscript, we have conducted additional experiments and provided quantitative data in response to the reviewers' comments. Furthermore, we have refined the manuscript and added further context to elucidate the significance of our findings for the readers.

Public Reviews:

Reviewer #1 (Public Review):

In this paper, Wu et al. investigated the physiological roles of CCDC113 in sperm flagellum and HTCA stabilization by using CRISPR/Cas knockouts mouse models, co-IP, and single sperm imaging. They find that CCDC113 localizes in the linker region among radial spokes, the nexin-dynein regulatory complex (N-DRC), and doublet microtubules (DMTs) RS, N-DRC, and DMTs and interacts with axoneme-associated proteins CFAP57 and CFAP91, acting as an adaptor protein that facilitates the linkage between RS, N-DRC, and DMTs within the sperm axoneme. They show the disruption of CCDC113 produced spermatozoa with disorganized sperm flagella and CFAP91, DRC2 could not colocalize with DMTs in Ccdc113-/- spermatozoa. Interestingly, the data also indicate that CCDC113 could localize on the HTCA region, and interact with HTCA-associated proteins. The knockout of Ccdc113 could also produce acephalic spermatozoa. By using Sun5 and Centlein knockout mouse models, the authors further find SUN5 and CENTLEIN are indispensable for the docking of CCDC113 to the implantation site on the sperm head. Overall, the experiments were designed properly and performed well to support the authors' observation in each part. Furthermore, the study's findings offer valuable insights into the physiological and developmental roles of CCDC113 in the male germ line, which can provide insight into impaired sperm development and male infertility. The conclusions of this paper are mostly well supported by data, but some points need to be clarified and discussed.

We thank Reviewer #1 for his or her critical reading and the positive assessment.

(1) In Figure 1, a sperm flagellum protein, which is far away from CCDC113, should be selected as a negative control to exclude artificial effects in co-IP experiments.

We greatly appreciate Reviewer #1’s insightful suggestion. In response, we selected two sperm outer dense fiber proteins, ODF1 and ODF2, which are located distant from the sperm axoneme, as negative controls in the co-IP experiments. As shown in Figure 1- figure supplement 1A and B, neither ODF1 nor ODF2 bound to CCDC113, indicating the interaction observed in Figure 1 is not an artifact.

(2) Whether the detachment of sperm head and tail in Ccdc113-/- mice is a secondary effect of the sperm flagellum defects? The author should discuss this point.

Good question. Considering that CCDC113 is localized in the sperm neck region and interacts with SUN5 and CENTLEIN, it may play a direct role in connecting the sperm head and tail. Indeed, PAS staining revealed that Ccdc113–/– sperm heads exhibit abnormal orientation in stages V–VIII of the seminiferous epithelia (Figure 6C-D). Furthermore, transmission electron microscopy (TEM) analysis indicated that the absence of CCDC113 caused detachment of the damaged coupling apparatus from the sperm head in step 9–11 spermatids (Figure 6E). These results suggest that the detachment of the sperm head and tail in Ccdc113–/– mice may not be a secondary effect of sperm flagellum defects. We have discussed this point further below:

“CCDC113 can interact with SUN5 and CENTLEIN, but not PMFBP1 (Figure 7A-C), and left on the tip of the decapitated tail in Sun5–/– and Centlein–/– spermatozoa (Figure 7K and L). Furthermore, CCDC113 colocalizes with SUN5 in the HTCA region, and immunofluorescence staining in spermatozoa shows that SUN5 is positioned closer to the sperm nucleus than CCDC113 (Figure 7G and H). Therefore, SUN5 and CENTLEIN may be closer to the sperm nucleus than CCDC113. PAS staining revealed that Ccdc113–/– sperm heads are abnormally oriented in stages V–VIII seminiferous epithelia (Figure6 C and D), and TEM analysis further demonstrated that the disruption of CCDC113 causes the detachment of the destroyed coupling apparatus from the sperm head in step 9–11 spermatids (Figure 6E). All these results suggest that the detachment of sperm head and tail in Ccdc113–/– mice may not be a secondary effect of sperm flagellum defects.”

(3) Given that some cytoplasm materials could be observed in Ccdc113-/- spermatozoa (Fig. 5A), whether CCDC113 is also essential for cytoplasmic removal?

Good question. Unremoved cytoplasm could be detected in spermatozoa by using transmission electron microscopy (TEM) analysis, including disrupted mitochondria, damaged axonemes, and large vacuoles. These observations indicate defects in cytoplasmic removal in Ccdc113–/– mice. We have discussed this point as below:

“Moreover, TEM analysis detected excess residual cytoplasm in spermatozoa, including disrupted mitochondria, damaged axonemes, and large vacuoles, indicating defects in cytoplasmic removal in Ccdc113–/– mice (Figure 5A).”

(4) Although CCDC113 could not bind to PMFBP1, the localization of CCDC113 in Pmfbp1-/- spermatozoa should be also detected to clarify the relationship between CCDC113 and SUN5-CENTLEIN-PMFBP1.

We appreciate Reviewer #1’s suggestion. We have analyzed the localization of CCDC113 in Pmfbp1-/- spermatozoa and found that CCDC113 was located at the tip of the decapitated tail in Pmfbp1-/- spermatozoa (Figure 7K and L). This finding has been incorporated into the revised manuscript as below:

“To further elucidate the functional relationships among CCDC113, SUN5, CENTLEIN, and PMFBP1 at the sperm HTCA, we examined the localization of CCDC113 in Sun5-/-, Centlein–/–, and Pmfbp1–/– spermatozoa. Compared to the control group, CCDC113 was predominantly localized on the decapitated flagellum in Sun5-/-, Centlein–/–, and Pmfnp1–/– spermatozoa (Figure 7K and L), indicating SUN5, CENTLEIN, and PMFBP1 are crucial for the proper docking of CCDC113 to the implantation site on the sperm head. Taken together, these data demonstrate that CCDC113 cooperates with SUN5 and CENTLEIN to stabilize the sperm HTCA and anchor the sperm head to the tail.”

Reviewer #2 (Public Review):

Summary:

In the present study, the authors select the coiled-coil protein CCDC113 and revealed its expression in the stages of spermatogenesis in the testis as well as in the different steps of spermiogenesis with expression also mapped in the different parts of the epididymis. Gene deletion led to male infertility in CRISPR-Cas9 KO mice and PAS staining showed defects mapped in the different stages of the seminiferous cycle and through the different steps of spermiogenesis. EM and IF with several markers of testis germ cells and spermatozoa in the epididymis indicated defects in flagella and head-to-tail coupling for flagella as well as acephaly. The authors' co-IP experiments of expressed CCDC113 in HEK293T cells indicated an association with CFAP91 and DRC2 as well as SUN5 and CENTLEIN.

The authors propose that CCDC113 connects CFAP91 and DRC2 to doublet microtubules of the axoneme and CCDC113's association with SUN5 and CENTLEIN to stabilize the sperm flagellum head-to-tail coupling apparatus. Extensive experiments mapping CCDC13 during postnatal development are reported as well as negative co-IP experiments and studies with SUN5 KO mice as well as CENTLEIN KO mice.

Strengths:

The authors provide compelling observations to indicate the relevance of CCDC113 to flagellum formation with potential protein partners. The data are relevant to sperm flagella formation and its coupling to the sperm head.

We are grateful to Reviewer #2 for his or her recognition of the strength of this study.

Weaknesses:

The authors' observations are consistent with the model proposed but the authors' conclusions for the mechanism may require direct demonstration in sperm flagella. The Walton et al paper shows human CCDC96/113 in cilia of human respiratory epithelia. An application of such methodology to the proteins indicated by Wu et al for the sperm axoneme and head-tail coupling apparatus is eagerly awaited as a follow-up study.

We thank Reviewer 2 for his/her kindly help in improving the manuscript. We now understand that directly detection of CCDC113 precise localization in sperm axoneme and head-tail coupling apparatus (HTCA) using cryo-electron microscopy (cryo-EM) could powerfully strengthen our model. Recent advances in cryo-EM have indeed advanced our understanding of axonemal structures analysis of axonemal structures and determined the structures of native axonemal DMTs from mouse, bovine, and human sperm (Leung et al., 2023; Zhou et al., 2023). However, high-resolution structures of sperm axoneme and HTCA regions, including those involving CCDC113, have yet to be fully characterized. Thus, we would like to discuss this point and consider it a valuable direction for future research.

“Given that the cryo-EM of sperm axoneme and HTCA could powerfully strengthen the role of CCDC113 in stabilizing sperm axoneme and head-tail coupling apparatus, it a valuable direction for future research.”

References:

Bazan, R., Schröfel, A., Joachimiak, E., Poprzeczko, M., Pigino, G., & Wloga, D. (2021). Ccdc113/Ccdc96 complex, a novel regulator of ciliary beating that connects radial spoke 3 to dynein g and the nexin link. PLoS Genet, 17(3), e1009388.

Ghanaeian, A., Majhi, S., McCafferty, C. L., Nami, B., Black, C. S., Yang, S. K., Legal, T., Papoulas, O., Janowska, M., Valente-Paterno, M., Marcotte, E. M., Wloga, D., & Bui, K. H. (2023). Integrated modeling of the Nexin-dynein regulatory complex reveals its regulatory mechanism. Nat Commun, 14(1), 5741.

Leung, M. R., Zeng, J., Wang, X., Roelofs, M. C., Huang, W., Zenezini Chiozzi, R., Hevler, J. F., Heck, A. J. R., Dutcher, S. K., Brown, A., Zhang, R., & Zeev-Ben-Mordehai, T. (2023). Structural specializations of the sperm tail. Cell, 186(13), 2880-2896.e2817

Walton, T., Gui, M., Velkova, S., Fassad, M. R., Hirst, R. A., Haarman, E., O'Callaghan, C., Bottier, M., Burgoyne, T., Mitchison, H. M., & Brown, A. (2023). Axonemal structures reveal mechanoregulatory and disease mechanisms. Nature, 618(7965), 625-633.

Zhou, L., Liu, H., Liu, S., Yang, X., Dong, Y., Pan, Y., Xiao, Z., Zheng, B., Sun, Y., Huang, P., Zhang, X., Hu, J., Sun, R., Feng, S., Zhu, Y., Liu, M., Gui, M., & Wu, J. (2023). Structures of sperm flagellar doublet microtubules expand the genetic spectrum of male infertility. Cell, 186(13), 2897-2910.e2819.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) Please provide full gel for the Figure 2C experiment (could be as a supplementary file).

Thanks for your insightful suggestions. We have replaced Figure 2C and provided the full gel in Figure 2-figure supplement 1A.

(2) The authors write on Line 163 "In contrast, the flagellum staining appeared reduced in Ccdc113-/- seminiferous tubules (Fig. 2J, red asterisk)." However, the magnification of the pictures is not sufficient to distinguish anything in the panel mentioned, please provide others.

Many thanks for pointing this out. We have provided the iconic figure to show the flagella defect in seminiferous tubules.

(3) Please add statistical p-values for figures.

Thanks for your valuable advice. We have added statistical p-values to the figures in the revised manuscript.

(4) Line 128: Should "speculate" be "speculated"?

Thank you for pointing out this problem. We have corrected it in the revised manuscript, as shown below:

“Given that CFAP91 has been reported to stabilize RS on the DMTs (Bicka et al., 2022; Dymek et al., 2011; Gui et al., 2021) and cryo-EM analysis shows that CCDC113 is closed to DMTs, we speculated that CCDC113 may connect RS to DMTs by binding to CFAP91 and microtubules.”

(5) In lines 384-385, more "-" is typed.

Thank you for pointing out this problem. We have corrected it in the revised manuscript, as shown below:

“Furthermore, CCDC113 colocalizes with SUN5 in the HTCA region, and immunofluorescence staining in spermatozoa shows that SUN5 is closer to the sperm nucleus than CCDC113 (Figure 7G and H). Therefore, SUN5 and CENTLEIN may be closer to the sperm nucleus than CCDC113.”

(6) In general, the article has many typos and should be professionally proofread.

Many thanks for pointing this out. We have thoroughly revised the manuscript with the assistance professional proofreading.

Reviewer #2 (Recommendations For The Authors):

Can the authors indicate in the Materials and Methods if n=3 biological replicates were done for all co-IP, EM, LM, and IF studies? The statistical analysis section indicates this but quantification is missing for most figures including co-IP, most IF, PAS staining, EM, etc.

We thank Reviewer 2 for the insightful comments and guidance to improve our data quality. All the experiments in this study were repeated at least three times to ensure reproducibility. We have quantified the co-IP experiments in Figures 1C-H and 7A-F, the IF data in Figures 2K, 5C, and 5D, as well as the PAS staining in Figure 6C. Since electron microscopy samples require very little testicular tissue and the sections obtained are very thin, the likelihood of capturing sections specifically at the sperm head-tail junction is considerably low. This challenge makes it difficult to perform quantitative analysis and statistical evaluation in the TEM experiment. To address this limitation, we have quantified the percentage of _Ccdc113-/-_sperm heads with abnormal orientation in stages V–VIII of the seminiferous epithelium to indicate impaired head-to-tail anchorage.

Figure S2 is compelling and might be indicated as a major figure instead of a supplementary figure.

We appreciate the positive comment. We have included it as a major figure in Figure 3F.

Figure 4A may be incomplete. Data sets for RNA expression suggest high expression in the ovary and other organs in males and females including the brain and are not indicated by the authors. Figure 4A may be considered for removal with a more complete study for another paper.

Thank you for pointing out this issue. We reviewed RNA expression data from various tissues using RNA-Seq data from Mouse ENCODE (https://www.ncbi.nlm.nih.gov/gene/244608) and found that CCDC113 is highly expressed in the testis, but not significantly in the ovary and brain (Figure 4- figure supplement 1A). Additionally, we re-evaluated CCDC113 protein levels in the spleen, lung, kidney, testis, intestine, stomach, brain, and ovary, confirming that it is highly expressed in the testes, with negligible expression in the ovary and brain (Figure 4- figure supplement 1B). In line with Reviewer 2's suggestion, we have removed Figure 4A in the revised manuscript.

There are grammatical errors throughout the manuscript and Figure 7 is truncated.

Thank you for pointing out this problem. We have thoroughly revised the manuscript with the assistance professional proofreading.

The Introduction and Discussion parts of the paper may need some clarification for the general reader. The material in the "Additional Context " section of the critique below may be a helpful place to introduce what a stage is, and the steps in germ cell development in the testis with the latter of course where and when the flagellum develops.

We appreciate your valuable suggestions. We have referred to the material in the “Additional Context” section to introduce the stages of spermatogenesis and the steps in germ cell development in the testis in the introduction and results.

“Male fertility relies on the continuous production of spermatozoa through a complex developmental process known as spermatogenesis. Spermatogenesis involves three primary stages: spermatogonia mitosis, spermatocyte meiosis, and spermiogenesis. During spermiogenesis, spermatids undergo complex differentiation processes to develop into spermatozoa, which includes nuclear elongation, chromatin remodeling, acrosome formation, cytoplasm elimination, and flagellum development (Hermo et al., 2010).”

Hermo, L., Pelletier, R. M., Cyr, D. G., & Smith, C. E. (2010). Surfing the wave, cycle, life history, and genes/proteins expressed by testicular germ cells. Part 1: background to spermatogenesis, spermatogonia, and spermatocytes. Microscopy research and technique, 73(4), 241–278. https://doi.org/10.1002/jemt.20783

“Pioneering work in the mid-1950s used the PAS stain in histologic sections of mouse testis to visualize glycoproteins of the acrosome and Golgi in seminiferous tubules (Oakberg, 1956). The pioneers discovered in cross-sectioned seminiferous tubules the association of differentiating germ cells with successive layers to define different stages that in mice are twelve, indicated as Roman numerals (XII). For each stage, different associations of maturing germ cells were always the same with early cells in differentiation at the periphery and more mature cells near the lumen. In this way, progressive differentiation from stem cells to mitotic, meiotic, acrosome-forming, and post-acrosome maturing spermatocytes was mapped to define spermatogenesis with the XII stages in mice representing the seminiferous cycle. The maturation process from acrosome-forming cells to mature spermatocytes is defined as spermiogenesis with 16 different steps that are morphologically distinct spermatids (O'Donnell L, 2015).”

Oakberg, E. F. (1956). A description of spermiogenesis in the mouse and its use in analysis of the cycle of the seminiferous epithelium and germ cell renewal. The American journal of anatomy, 99(3), 391-413. https://doi.org/10.1002/aja.1000990303

O'Donnell L. (2015). Mechanisms of spermiogenesis and spermiation and how they are disturbed. Spermatogenesis, 4(2), e979623. https://doi.org/10.4161/21565562.2014.979623

For the Discussion, the authors indicate that the function of CCDC113 in mammals is unknown yet the authors point to the work of Walton et al on human respiratory epithelia that points to a function for CCDC96/113. The work in the manuscript here does indicate a role in sperm flagella and the head-to-tail coupling apparatus but remains descriptive until the methodology of Walton et al is applied. Hopefully, the authors will consider it for a follow-up study.

Thank you for pointing out this problem. We have revised this part and highlighted the Walton et al’s work in the Discussion.

“CCDC113 is a highly evolutionarily conserved component of motile cilia/ﬂagella. Studies in the model organism, Tetrahymena thermophila, have revealed that CCDC113 connects RS3 to dynein g and the N-DRC, which plays essential role in cilia motility (Bazan et al., 2021; Ghanaeian et al., 2023). Recent studies have also identified the localization of CCDC113 within the 96-nm repeat structure of the human respiratory epithelial axoneme, and localizes to the linker region among RS, N-DRC and DMTs (Walton et al., 2023). In this study, we reveal that CCDC113 is indispensable for male fertility, as Ccdc113 knockout mice produce spermatozoa with flagellar defects and head-tail linkage detachment (Figure 3D).”

“Overall, we identified CCDC113 as a structural component of both the flagellar axoneme and the HTCA, where it performs dual roles in stabilizing the sperm axonemal structure and maintaining the structural integrity of HTCA. Given that the cryo-EM of sperm axoneme and HTCA could powerfully strengthen the role of CCDC113 in stabilizing sperm axoneme and head-tail coupling apparatus, it a valuable direction for future research.”

The Discussion may be focused on the key aspects of CCDC113 related to sperm flagella and the head-to-tail coupling apparatus that represent a genuine advance. The more speculative parts of the Discussion that have not been addressed by experimentation in the Results section may be considered for removal in the Discussion section.

Thank you for pointing out this. We have removed the speculative parts of the Discussion that have not been addressed by experimentation in the Results section.

Additional Context to help readers understand the significance of the work:

Pioneering work in the mid-1950s used the periodic acid Schiff (PAS) stain in histologic sections of rodent testis to visualize glycoproteins of the acrosome and Golgi in seminiferous tubules. The pioneers discovered in cross-sectioned seminiferous tubules the association of differentiating germ cells with successive layers to define different stages that in mice are twelve, indicated as Roman numerals (XII). For each stage, different associations of maturing germ cells were always the same with early cells in differentiation at the periphery and more mature cells near the lumen. In this way, progressive differentiation from stem cells to mitotic, meiotic, acrosome-forming, and post-acrosome maturing spermatocytes was mapped to define spermatogenesis with the XII stages in mice representing the seminiferous cycle. The maturation process from acrosome-forming cells to mature spermatocytes is defined as spermiogenesis with 19 different steps that are morphologically distinct spermatids. It is from steps 8-19 of spermiogenesis that the formation of the flagellum takes place. Final maturation occurs in the epididymis as sperm move through the caput, corpus, and cauda of the organ with motile spermatozoa generated.

Thank you very much!

Version 2

Abstract

Abstract

The structural integrity of the sperm is crucial for male fertility, defects in sperm head-tail linkage and flagellar axoneme are associated with acephalic spermatozoa syndrome (ASS) and the multiple morphological abnormalities of the sperm flagella (MMAF). Notably, impaired head-tail coupling apparatus (HTCA) often accompanies defects in the flagellum structure, however, the molecular mechanisms underlying this phenomenon remain elusive. Here, we identified an evolutionarily conserved coiled-coil domain-containing (CCDC) protein, CCDC113, and found the disruption of CCDC113 produced spermatozoa with disorganized sperm flagella and HTCA, which caused male infertility. Further analysis revealed that CCDC113 could bind to CFAP57 and CFAP91, and function as an adaptor protein for the connection of radial spokes, nexin-dynein regulatory complex (N-DRC) and doublet microtubules (DMTs) in the sperm axoneme. Moreover, CCDC113 was identified as a structural component of HTCA, collaborating with SUN5 and CENTLEIN to connect sperm head to tail during spermiogenesis. Together, our studies reveal that CCDC113 serve as a critical hub for sperm axoneme and HTCA stabilization, providing insights into the potential pathogenesis of infertility associated with human CCDC113 mutations.

Introduction

Male fertility relies on the continuous production of spermatozoa through a complex developmental process known as spermatogenesis. Spermatogenesis involves three primary stages: spermatogonia mitosis, spermatocyte meiosis, and spermiogenesis. During spermiogenesis, spermatids undergo complex differentiation processes to develop into spermatozoa, which includes nuclear elongation, chromatin remodeling, acrosome formation, cytoplasm elimination, and flagellum development (Hermo et al., 2010). The integrity of spermatozoa is essential for their migration through the female reproductive tract and subsequent successful fertilization. An intact spermatozoon contains properly formed sperm head, head-tail coupling apparatus (HTCA), and flagellum (Parker, 2020; Roldan, 2019). Numerous components have been identified in the sperm HTCA and flagellum that are essential for the sperm integrity (Lehti & Sironen, 2017; Wu et al., 2020). Defects in sperm flagellum and HTCA can lead to reduced sperm motility or abnormal sperm morphology, termed as multiple morphological abnormalities of the sperm flagella (MMAF) or acephalic spermatozoa syndrome (ASS), which, in turn, causes male infertility (Sudhakar et al., 2021; Tu et al., 2020).

The sperm flagellum possesses an evolutionarily conserved axonemal structure composed of “9+2” microtubules, specifically, nine peripheral doublet microtubules (DMTs) surrounding two central microtubules known as the central pair (CP) (Inaba & Mizuno, 2016). Axonemal dyneins, radial spokes (RS), and the nexin-dynein regulatory complex (N-DRC) are arranged on DMTs with a 96-nm repeating unit structures (Kumar & Singh, 2021). Within the axoneme, the N-DRC and RS are crucial for maintaining axonemal integrity, forming crossbridges between adjacent DMTs and linking the DMTs to the central apparatus, respectively (Canty et al., 2021; Ishikawa, 2017; Kumar & Singh, 2021). Recent advancements in artificial intelligence, biochemical techniques, and cryo-electron microscopy (cryo-EM) facilitated the analysis of the axonemal structures, revealing numerous components among RS, N-DRC and DMTs that may serve as hubs for axoneme stabilization (Bazan et al., 2021; Leung et al., 2023; Walton et al., 2023; Zhou et al., 2023). For example, CFAP91 has been identified as a protein that extends from the base of RS2 through the N-DRC base plate to RS3, thus stabilizing RS2 and RS3 on the DMTs (Bicka et al., 2022; Dymek et al., 2011; Gui et al., 2021). Similarly, CFAP57 extends through the N-DRC and interacts with RS3 via its C-terminal region (Ghanaeian et al., 2023). Recent analyses suggest that CCDC96 and CCDC113 may form a complex that extends parallel to the N-DRC, connecting the base of RS3 to the tail of dynein g (IDA g) and the N-DRC (Bazan et al., 2021; Ghanaeian et al., 2023). However, the functions of these proteins in stabilizing the sperm flagellum remain unknown.

The sperm flagellum is tightly anchored to the sperm head through the HTCA, a complex structure based on the centrosome (Wu et al., 2020). This structure consists of two cylindrical microtubule-based centrioles and associated components, including well-organized segmented columns, capitulum plate and basal plate. The segmented columns and capitulum plate, located below the basal plate, are thought to originate from the dense material emanating from the proximal centriole (Fawcett & Phillips, 1969; Zamboni & Stefanini, 1971). Many proteins have been identified in the sperm HTCA, with mouse models exhibiting phenotypes characteristic of acephalic spermatozoa syndrome (Wu et al., 2020). SPATA6 is the first protein identified as a component of the HTCA using the knockout mouse model, and is crucial for its formation (Yuan et al., 2015). Deficiencies in SUN5 (Elkhatib et al., 2017; Fang et al., 2018; Liu et al., 2020; Sha et al., 2018; Shang et al., 2018; Shang et al., 2017; Xiang et al., 2022; Zhang et al., 2021; Zhu et al., 2016) and PMFBP1 (Deng et al., 2022; Liu et al., 2020; Liu et al., 2021; Lu et al., 2021; Nie et al., 2022; Sha et al., 2019; Zhu et al., 2018) have been associated with ASS in both humans and mice. The centriole-related protein CENTLEIN acts as a bona fide linker between SUN5 and PMFBP1, participating in the HTCA assembly (Zhang et al., 2021). Notably, impaired HTCA often coincides with defects in the sperm flagellum (Hall et al., 2013; Shang et al., 2017; Yuan et al., 2015; Zhang et al., 2021; Zhu et al., 2018), suggesting that the stabilization of the HTCA may be closely associated with the integrity of sperm flagellum. However, the mechanism that maintains the stabilization of both the sperm flagellum and the HTCA remains to be clarified.

Here, we identified an evolutionarily conserved coiled-coil domain-containing (CCDC) protein, CCDC113, and found that it forms a complex with CFAP57 and CFAP91, thereby facilitating the connection of RS, N-DRC and DMTs in the axoneme. Knockout of Ccdc113 resulted in spermatozoa with flagellar defects and head-tail linkage detachment, leading to male infertility. Ultrastructural analysis showed that the loss of CCDC113 disrupted both the sperm axoneme and HTCA. CCDC113 localizes on the manchette, HTCA and flagellum in elongating and elongated spermatids. Further analysis revealed that CCDC113 is indispensable for the connection of CFAP91 and DRC2 with DMTs in the sperm axoneme, and it interacts with SUN5 and CENTLEIN to stabilize the sperm HTCA. These results suggest that CCDC113 serves as a critical hub in maintaining the structural integrity of both the sperm flagellum and HTCA.

Results

CCDC113 complexes with CFAP57 and CFAP91

CCDC113 is an evolutionarily conserved coiled-coil domain-containing (CCDC) protein identified in the ciliated species. Comparative analysis of CCDC113 structures from Tetrahymena thermophila to Homo sapiens showed structural similarity among CCDC113 orthologs (Figure 1A). Recent cryo-EM analysis in the structure of the 96-nm modular repeats of axonemes from the Tetrahymena thermophila cilia and human respiratory cilia revealed that CCDC113 localizes to the linker region among RS, N-DRC and DMTs (Figure 1B), suggesting it may serve as a structural component connecting RS, N-DRC and DMTs (Bazan et al., 2021; Ghanaeian et al., 2023). To further investigate, we examined the interactions between CCDC113 and its neighboring axoneme-associated proteins, CFAP57 and CFAP91 (Figure 1B). We transfected HEK293T cells with a GFP-tagged CCDC113 and FLAG-tagged CFAP57 or CFAP91, then performed anti-FLAG-immunoprecipitations. CCDC113 was present in both FLAG-CFAP57 and FLAG-CFAP91 immunoprecipitates (Figure 1C and D), indicating CCDC113 interacts with both CFAP57 and CFAP91. Given that CFAP91 has been reported to stabilize RS on the DMTs (Bicka et al., 2022; Dymek et al., 2011; Gui et al., 2021) and cryo-EM analysis shows that CCDC113 is closed to DMTs, we speculated that CCDC113 may connect RS to DMTs by binding to CFAP91 and microtubules. To test this, we detected the interaction between CCDC113 and β-tubulin (TUBB5) and found that CCDC113 was present in MYC-TUBB5 immunoprecipitate (Figure 1E). Additionally, since CFAP57 extends through the N-DRC and CCDC113 is closed to the N-DRC (Ghanaeian et al., 2023), we further examined the interaction between CCDC113 and N-DRC components adjacent to DMTs. Co-immunoprecipitation (co-IP) analysis showed that CCDC113 could bind to DRC1, DRC2, and DRC3 (Figure 1F-H). We also included two sperm outer dense fiber proteins, ODF1 and ODF2 (Zhu et al., 2022), which are located far from the sperm axoneme, as negative controls in the co-IP experiments. As shown in Figure 1-figure supplement 1A and 1B, neither ODF1 nor ODF2 bound to CCDC113. Therefore, CCDC113 may function as an adaptor protein connecting RS, N-DRC, and DMTs, serving as a critical hub for axoneme stabilization.

CCDC113 is an evolutionarily conserved axoneme-associated protein.

(A) Multiple species phylogenetic tree of CCDC113. Structural similarity scores (Z scores) of CCDC113 orthologs in Homo sapiens, Mus musculus, Phascolarctos cinereus, Danio rerio, Chlamydomonas reinhardtii and Tetrahymena thermophila were derived through the DALI webserver for pairwise structure comparisons (Holm & Laakso, 2016). (B) Positioning of CCDC113 within the 96-nm repeat of human axoneme (Walton et al., 2023). CCDC113 forms a complex with CCDC96, is located at the base of RS3, and is adjacent to CFAP91 and CFAP57. CFAP91 originates at the base of RS2 and links the RS3 subunits (CFAP251 and CFAP61). (C-H) Neighboring axoneme-associated proteins were expressed alone or co-expressed with CCDC113 in HEK293T cells, and the interactions between CCDC113 and CFAP57, CFAP91, TUBB5, DRC1, DRC2, or DRC3 were examined by co-immunoprecipitation. IB: immunoblotting; IP: immunoprecipitation. The % Input is displayed below the corresponding figures for quantification. n = 3 independent experiments. Data are presented as mean ± SD; *P < 0.05, **P < 0.01, ***P < 0.001.

CCDC113 is required for male fertility

To investigate the physiological functions of CCDC113, we generated a Ccdc113 knockout mouse strain using the CRISPR/Cas9 system (Figure 2A). The Ccdc113−/−mice were genotyped by genomic DNA sequencing and further confirmed by polymerase chain reaction (PCR). Genotypes were distinguished by a 539 bp band for Ccdc113+/+ mice, a 461 bp band for Ccdc113−/−mice, and two bands of 539 bp and 461 bp for Ccdc113+/− mice (Figure 2B). Further immunoblotting analysis confirmed the elimination of CCDC113 in total protein extracts from Ccdc113−/− testes (Figure 2C; Figure 2-figure supplement 1A). Mice lacking Ccdc113 showed no gross abnormalities in their appearance or behavior, and no obvious differences in body weight (Figure 2H). Additionally, no hydrocephalus or left-right asymmetry defects were observed (Figure 2-figure supplement 1B). Additionally, the deficiency of CCDC113 did not affect ciliogenesis in the lung and trachea (Figure 2-figure supplement 1C-E). We then assessed the fertility of 2-month-old male and female Ccdc113−/− mice. Ccdc113−/− female mice were able to produce offspring after mating with wild-type (WT) adult males, similar to Ccdc113+/+ female mice (Figure 2D). However, Ccdc113−/− male mice exhibited normal mating behavior, as indicated by the presence of copulatory plugs, they failed to produce offspring when mating with WT adult female mice (Figure 2E). Thus, the knockout of Ccdc113 results in male infertility.

Ccdc113 knockout leads to male infertility.

(A) The CRISPR-Cas9 strategy for generating the Ccdc113 knockout mice. (B) Genotyping to identify Ccdc113 knockout mice. (C) Immunoblotting of CCDC113 in Ccdc113+/+ and Ccdc113−/− testes. TUBULIN served as the loading control. (D) The average litter size of Ccdc113+/+ and Ccdc113−/− female mice in 2 months (n = 5 independent experiments). Data are presented as mean ± SD; ns indicates no significant difference. (E) The average litter size of Ccdc113+/+ and Ccdc113−/− male mice in 2 months (n = 5 independent experiments). Data are presented as mean ± SD; ****P < 0.0001. (F) The size of testes was similar in Ccdc113+/+ and Ccdc113−/− mice. (G) The testis weights of Ccdc113+/+ and Ccdc113−/− male mice (n = 5 independent experiments). Data are presented as mean ± SD; ns indicates no significant difference. (H) The body weights of Ccdc113+/+ and Ccdc113−/− male mice (n = 5 independent experiments). Data are presented as mean ± SD; ns indicates no significant difference. (I) The ratio of testis weight/body weight in Ccdc113+/+ and Ccdc113−/− male mice (n = 5 independent experiments). Data are presented as mean ± SD; ns indicates no significant difference. (J) H&E staining of testis sections from Ccdc113+/+ and Ccdc113−/− male mice. Red asterisks indicate the abnormal sperm flagellum in the Ccdc113−/− testis seminiferous tubule. (K) Immunofluorescence of acetylated-tubulin (red) in testis sections from Ccdc113−/− male mice showed flagellar defects. (L) Acetylated-tubulin fluorescence intensity was measured per testis section in 155 sections from 3 Ccdc113+/+ mice and 153 sections from 3 Ccdc113−/−male mice. Data are presented as mean ± SD; ****P < 0.0001.

Ccdc113 knockout mice produce spermatozoa with flagellar defects and head-tail linkage detachment

To further investigate the cause of male infertility, we initially examined Ccdc113-/- testis at both macroscopic and histological levels. Ccdc113 knockout did not affect testis size (Figure 2F and G) or the ratio of testis weight to body weight (Figure 2H and I). Histological sections stained with hematoxylin-eosin (H&E) revealed that seminiferous tubules of Ccdc113+/+ mice exhibited a tubular lumen with flagella emerging from the developing spermatids. In contrast, flagellar staining appeared reduced in Ccdc113−/− seminiferous tubules (Figure 2J, red asterisks). Immunofluorescence staining for acetylated tubulin (ac-tubulin), a marker for sperm flagellum (Martinez et al., 2020), further confirmed the flagellar defects in Ccdc113−/− mice (Figure 2K and L).

Subsequently, we examined spermatids at different stages in Ccdc113−/− testes using periodic acid Schiff (PAS) staining. Pioneering work in the mid-1950s used the PAS stain in histologic sections of mouse testis to visualize glycoproteins of the acrosome and Golgi in seminiferous tubules (Oakberg, 1956). The pioneers discovered in cross-sectioned seminiferous tubules the association of differentiating germ cells with successive layers to define different stages that in mice are twelve, indicated as Roman numerals (XII). For each stage, different associations of maturing germ cells were always the same with early cells in differentiation at the periphery and more mature cells near the lumen. In this way, progressive differentiation from stem cells to mitotic, meiotic, acrosome-forming, and post-acrosome maturing spermatocytes was mapped to define spermatogenesis with the XII stages in mice representing the seminiferous cycle. The maturation process from acrosome-forming cells to mature spermatocytes is defined as spermiogenesis with 16 different steps that are morphologically distinct spermatids (O’Donnell L, 2015). While acrosome biogenesis and nuclear morphology in Ccdc113−/− spermatids from steps 1 to 10 were comparable to those in Ccdc113+/+ spermatids, abnormal club-shaped heads were observed in spermatids from steps 11 to 16 in Ccdc113−/− mice (Figure 3-figure supplement 1A, black asterisk). In addition, the manchette of Ccdc113−/−spermatids was more elongated compared to that of Ccdc113+/+ spermatids (Figure 3-figure supplement 1B). Therefore, the disruption of CCDC113 impaired spermiogenesis.

Ccdc113 knockout results in sperm flagellar defects and sperm head-tail detachment.

(A) H&E staining of the cauda epididymis from 2-month-old Ccdc113+/+and Ccdc113−/− male mice. Red circles indicate coiled eosin-stained structures without sperm heads in the epididymal lumen. (B) Analysis of sperm counts in Ccdc113+/+ and Ccdc113−/−male mice (n = 5 independent experiments). Mature spermatozoa were extracted from the unilateral cauda epididymis and dispersed in PBS. Sperm counts were measured using hemocytometers. Data are presented as mean ± SD; ****P < 0.0001. (C) Motile sperm in Ccdc113+/+ and Ccdc113-/- mice (n = 5 independent experiments). Data are presented as mean ± SD; ****P < 0.0001. (D) Ccdc113+/+ and Ccdc113−/− spermatozoa were co-stained with a flagellar marker α/β-tubulin (red) and an acrosomal marker PNA. Nuclei were stained with DAPI (blue). (E) Quantification of different categories of Ccdc113+/+, Ccdc113−/−spermatozoa (n = 3 independent experiments). Data are presented as mean ± SD. (F) Immunofluorescence analysis of acetylated-tubulin (green) and PNA (red) from Ccdc113+/+ and Ccdc113-/- spermatids.

Next, we examined the spermatozoa in the cauda epididymis and found that the sperm count in the Ccdc113−/− cauda epididymis was significantly decreased compared to the control group (Figure 3A and B). The motility of the released spermatozoa from Ccdc113+/+ and Ccdc113−/− cauda epididymis showed that Ccdc113−/− spermatozoa were completely immotile (Figure 3C). H&E staining of the cauda epididymis showed fewer hematoxylin-stained sperm heads in the Ccdc113−/− cauda epididymis compared to the Ccdc113+/+ cauda epididymis. Notably, unlike the control group, which exhibited linear eosin staining in the epididymal lumen, the Ccdc113−/− mice showed numerous coiled eosin-stained structures without sperm heads in the epididymal lumen (Figure 3A, red circles). To determine the morphological characteristics of the spermatozoa, we conducted single-sperm immunofluorescence using an anti-α/β-tubulin antibody to label the sperm flagellum and lectin peanut agglutinin (PNA) to visualize the sperm acrosome (Nakata et al., 2015). We noticed that Ccdc113−/− spermatozoa showed severe morphological malformations, including sperm head-tail detachment (type 1), abnormal sperm head with curly tail (type 2), normal sperm head with curly tail (type 3) (Figure 3D and E). To determine the role of CCDC113 in flagellum assembly, we analyzed flagellum formation during spermiogenesis. We found that in Ccdc113−/− mice, diffuse and curly axonemal signals were observed in testicular germ cells starting from the round spermatid stage (Figure 3F). Overall, these findings suggest that the deletion of CCDC113 leads to sperm flagellum deformities and detachment of the sperm head-to-tail linkage, resulting in a unique type of acephalic spermatozoa, which may be responsible for the Ccdc113−/− male infertility.

CCDC113 localizes on the sperm neck and flagellum regions

To gain further insights into the functional role of CCDC113 during spermiogenesis, we examined its expression in different tissues and found that CCDC113 was predominantly expressed in mouse testis (Figure 4-figure supplement 1A and B). CCDC113 was first detected in testis at postnatal day 7 (P7), with expression levels increasing continuously from P21 onward, peaking in adult testes (Figure 4A), suggesting that CCDC113 is highly expressed throughout spermiogenesis. We then conducted immunofluorescence analysis of CCDC113 in Ccdc113+/+ and Ccdc113−/− germ cells to determine its precise localization during spermatogenesis. CCDC113 appeared as punctate signals near the nuclei of spermatocyte and round spermatids, and localized to the manchette, sperm neck and flagellum regions in elongating and elongated spermatids (Figure 4-figure supplement 1C).

CCDC113 localizes to the HTCA, manchette and sperm flagellum.

(A) CCDC113 was expressed starting in P7 testes. TUBULIN served as the loading control. An asterisk indicates nonspecific bands. (B) Immunofluorescence of CCDC113 (red) and CENTRIN1/2 (green) in developing germ cells. CCDC113 partially colocalize with centriolar protein CENTRIN1/2. (C) Immunofluorescence of CCDC113 (red) and α/β-tubulin (green) in developing germ cells. The manchette was stained with the anti-α/β-tubulin antibody. (D-E) CCDC113 localizes to the HTCA and flagellum in mature mouse spermatozoa and human spermatozoa. Nuclei were stained with DAPI (blue).

To further validate these findings, we co-stained CCDC113 with α/β-tubulin, which marks the manchette and flagellum in spermatids (Lehti & Sironen, 2016) (Figure 4B). The immunofluorescence analysis showed that CCDC113 localized to the manchette surrounding the spermatid head from step 9 to step 14, as well as to the sperm neck and flagellum in the testes (Figure 4B). Given that CCDC113 was initially identified as a component of centriolar satellites (Firat-Karalar et al., 2014), the punctate signals of CCDC113 in spermatocyte and spermatids may be localized around the centrosome. To test this, we performed the immunofluorescent staining of CCDC113 and centrosomal protein CENTRIN1/2 in spermatocytes and spermatids, and found that the signal of CCDC113 partially colocalized with CENTRIN1/2 (Figure 4C). Thus, CCDC113 is localized to the centrosome, manchette, sperm neck and flagellum regions in the developing germ cells.

Next, we examined the localization of CCDC113 in mature spermatozoa, and found that CCDC113 was localized in the sperm neck and flagellum regions (Figure 4D). Similar localization of CCDC113 was also observed in human mature spermatozoa (Figure 4E). The consistent localization of CCDC113 at the sperm neck and flagellum in mature spermatozoa suggests its importance for maintaining the integrity of the sperm flagellum and head-to-tail connection.

Ccdc113 knockout results in the disorganization of the sperm flagellum structures

To delineate the sperm flagellar defects in Ccdc113−/− mice, we conducted transmission electron microscopy (TEM) examination of longitudinal sections of Ccdc113−/− spermatozoa. TEM analysis revealed a significant presence of unremoved cytoplasm, including disrupted mitochondria, damaged axonemes, and large vacuoles in Ccdc113−/− spermatozoa (Figure 5A, red asterisks). Cross sections of the principal piece of Ccdc113−/− spermatozoa further revealed partial loss or unidentifiable “9+2” axonemal structures, along with the disruption of the fibrous sheath and outer dense fibers (Figure 5A). We further examined the axonemal structure in Ccdc113−/− testicular spermatids using TEM and found disorganized axonemal microtubules were detected in Ccdc113−/− testicular spermatids (Figure 5B). In contrast to the regular positioning of the CP and nine peripheral DMTs in the Ccdc113+/+ spermatid axoneme, the Ccdc113−/− spermatids exhibited a scattered arrangement of DMTs, and no distinct radial spokes were observed (Figure 5B, red arrowheads). These results indicate CCDC113 is essential for the integrity of sperm flagellum.

CCDC113 is indispensable for the docking of CFAP91 and DRC2 to the DMTs to maintain the structural integrity of the axoneme.

(A) TEM analysis of spermatozoa from the cauda epididymidis of Ccdc113+/+ and Ccdc113−/− male mice. The flagellar longitudinal sections of Ccdc113−/− spermatozoa revealed unremoved cytoplasm, including disrupted mitochondria, damaged axonemes. and large vacuoles. Asterisks indicate large vacuoles. Cross sections of the principal piece of Ccdc113−/− spermatozoa further revealed partial loss or unidentifiable “9+2” structures, along with the disruption of the fibrous sheath and outer dense fibers. (B) TEM analysis of the axoneme in testicular spermatids from Ccdc113+/+ and Ccdc113−/− male mice. The red arrowheads indicate the absence of significant radial spokes (RSs). MS: mitochondrial sheath, Mi: mitochondrial, AX: axoneme. FS: fibrous sheath. DMT: doublet microtubule, MT: microtubule, CP: central pair, ODF: outer dense fiber, RS: radial spokes. (C) The immunofluorescence analysis for CFAP91 (green) and α/β-tubulin (red) was performed in Ccdc113+/+ and Ccdc113−/− spermatozoa. Nuclei were stained with DAPI (blue). White asterisks indicate regions not co-localized with tubulin. (D) The immunofluorescence analysis for DRC2 (green) and α/β-tubulin (red) was performed in Ccdc113+/+ and Ccdc113−/− spermatozoa. Nuclei were stained with DAPI (blue). White asterisks indicate regions not co-localized with tubulin. (E, G) Line-scan analysis (white line) was performed using Image J software. (F, H) Percentage of abnormally distributed CFAP91 and DRC2 on the axoneme of Ccdc113+/+ and Ccdc113−/− spermatozoa (n = 3 independent experiments). At least 200 spermatozoa were analyzed from each mouse. Data are presented as mean ± SD; ****P < 0.0001.

CCDC113 has been shown to localize at the base of the RS3 and interact with adjacent axoneme-associated proteins (Figure 1B-H). Given the disorganized “9+2” axonemal structure was detected in cross-sectioned Ccdc113−/− flagellar specimen (Figure5A and B), we speculated that CCDC113 likely served as an adaptor to connect the neighboring axoneme-associated proteins to DMTs. To test this, we examined the flagellar localization of CFAP91 in Ccdc113−/− spermatozoa, which is positioned in close proximity to CCDC113 at the root region of RS3 and is critical for the localizations of calmodulin-associated and spoke-associated complex (CSC) proteins CFAP61 and CFAP251 (Bicka et al., 2022; Meng et al., 2024). Immunofluorescence results indicated that the absence of Ccdc113 leads to the abnormal distribution of CFAP91 on the axoneme, where CFAP91 could not colocalize with DMTs. (Figure 5C, white asterisks, E and F). Given that DRC2 serves as the core component of the axonemal N-DRC (Jreijiri et al., 2023) and CCDC113 could bind to DRC2 (Figure 1G), we further examined DRC2 localization in the Ccdc113+/+ and Ccdc113−/− spermatozoa.

Immunofluorescence analysis showed that DRC2 exhibited distinct signals that did not colocalize with DMTs of Ccdc113−/− spermatozoa (Figure 5D, white asterisks, G and H). These findings collectively indicate that CCDC113 is indispensable for the connection of CFAP91 and DRC2 to the DMTs, which is required for structural integrity of the sperm axoneme.

Ccdc113 knockout impairs head-to-tail anchorage of the spermatids

To explore the mechanism behind acephalic spermatozoa in Ccdc113 knockout mice, we first examined where the flagellum detached from sperm head in Ccdc113−/− mice. The proportion of decapitated tails in the caput, corpus and cauda of Ccdc113−/− epididymis was similar (Figure 6A), suggesting the separation of the sperm head and tail in Ccdc113−/−mice may occur either within the seminiferous tubules or upon entering the caput of the epididymis. To confirm this, we performed PAS staining to examine spermiogenesis stages in Ccdc113+/+ and Ccdc113−/− testes (Figure 6B). We found that in Ccdc113+/+ testes, sperm heads at stages VII-VIII were oriented towards the basal membrane, whereas in Ccdc113−/− testes, sperm heads were oriented towards the tubule lumen during these stages (Figure 6C, arrows and D), which may be due to the separation of sperm heads from the flagellum during spermiogenesis. Additionally, mature sperm heads were still present at stages IX–X in Ccdc113−/− testes, whereas mature spermatozoa were released into the lumen of the seminiferous tubule at stage VIII in Ccdc113+/+ testes (Figure 6B, red asterisk). These results suggest that the sperm head and flagellum separation may occur during spermiation in the Ccdc113−/− testes.

Ccdc113 knockout spermatids display impaired HTCA.

(A) The proportion of decapitated tails in Ccdc113+/+ and Ccdc113−/−corpus, caput and cauda epididymis (n = 3 independent experiments). At least 200 spermatozoa were analyzed from each mouse. Data are presented as mean ± SD; ****P < 0.0001, ***P < 0.001. (B) PAS staining of testis sections from Ccdc113+/+and Ccdc113−/− mice. The green arrows indicate the orientation of the sperm heads in stages V–VIII seminiferous epithelia. Ccdc113−/− sperm head could still be detected in stages IX–X seminiferous epithelia. P: pachytene spermatocyte, spz: spermatozoa, rSt: round spermatid, eSt: elongating spermatid, Z: zygotene spermatocyte, M: meiotic spermatocyte. (C) Ccdc113−/− spermatids lost their head orientation toward the basement membrane during spermiation in stages VII–VIII of the seminiferous epithelium. L: lumen, B: basement membrane. (D) Percentage of sperm heads with abnormal orientation in stages VII– VIII of the seminiferous epithelium in Ccdc113+/+ and Ccdc113−/− mice (n = 3 independent experiments). At least 200 spermatozoa were analyzed from each mouse. Data are presented as mean ± SD; ****P < 0.0001. (E) Defective HTCA formation in Ccdc113−/− spermatids. TEM analysis of the stepwise development of the HTCA were performed in Ccdc113+/+ and Ccdc113−/− testes. In Ccdc113+/+ spermatids, the well-defined coupling apparatus was tightly attached to the sperm head. In Ccdc113−/−spermatids, segmented columns (Scs), the capitulum (Cp) were absent. The red asterisks indicate the distance between the sperm head and HTCA. The white arrows indicate the dense material surrounding the proximal centriole. Nu: nuclear, Bp: basal plate, Cp: capitulum, Sc: segmented column, Pc: proximal centriole, Dc: distal centriole, An: annulus, Ax: axoneme, Rn: redundant nuclear envelope.

Next, we examined the development of the HTCA in Ccdc113+/+and Ccdc113−/− spermatids using TEM. In Ccdc113+/+ step 9–11 spermatids, the well-defined coupling apparatus, comprising the basal plate, capitulum plate, segmented columns, proximal centriole, distal centriole, was tightly attached to the sperm head. However, in Ccdc113−/− step 9–11 spermatids, the abnormal HTCA was detached from the sperm head (Figure 6E, red asterisk). Further observation of the HTCA structure revealed the absence of segmented columns and capitulum plate; only dense material surrounding the proximal centriole and basal plate could be detected (Figure 6E, white arrow). The basal plates were abnormally distant from their native implantation site on the nucleus of Ccdc113−/− elongating and elongated spermatids (Figure 6E). Taken together, our results indicate that the disruption of Ccdc113 causes the destroyed coupling apparatus detachment from the sperm head during spermiogenesis, and CCDC113 is required for the integrity of the sperm HTCA.

CCDC113 cooperates with SUN5 and CENTLEIN to stabilize sperm HTCA

To elucidate the molecular function of CCDC113 in sperm head-tail linkage, we examined its interaction with known HTCA-associated proteins, including SUN5, CENTLEIN, PMFBP1, and SPATA6 (Shang et al., 2017; Yuan et al., 2015; Zhang et al., 2021; Zhu et al., 2018). GFP-tagged CCDC113 and FLAG-tagged HTCA-associated proteins were co-transfected into HEK293T cells, followed by the immunoprecipitation with anti-GFP antibody. We found that CCDC113 interacted with SUN5 and CENTLEIN, but not with PMFBP1 and SPATA6 (Figure 7A-C and F). We further conducted co-IP assays in the reverse direction and verified that both SUN5 and CENTLEIN could bind to CCDC113 (Figure 7D and E). To investigate the localization of CCDC113 within the HTCA, we co-stained mature spermatozoa with antibodies against CCDC113 and SUN5, which has been reported to be localized at the root connecting the HTCA to the nuclear envelope (Shang et al., 2017; Zhang et al., 2021). CCDC113 was found positioned below SUN5, showed partial overlap with it (Figure 7G). The centriolar protein CENTLEIN was localized to the HTCA and served as the critical linker protein between SUN5 and PMFBP1 in the elongating and elongated spermatid (Zhang et al., 2021)). Given that CENTLEIN disappears in mature spermatozoa (Zhang et al., 2021), we performed the immunofluorescent staining of CCDC113 and CENTLEIN in elongated spermatids and found that CCDC113 partially colocalized with CENTLEIN at the HTCA (Figure 7H). Thus, CCDC113 interacts with both SUN5 and CENTLEIN and localizes at the sperm HTCA.

CCDC113 interact with SUN5 and CENTLEIN, participating in sperm head-tail linkage.

(A-C, F) HTCA-associated proteins (SUN5, CENTLEIN, PMFBP1, SPATA6) were expressed alone or co-expressed with CCDC113 in HEK293T cells, and the interactions between CCDC113 and these HTCA-associated proteins were examined by co-immunoprecipitation. CCDC113 interacted with SUN5 and CENTLEIN, but did not interact with PMFBP1 and SPATA6. IB: immunoblotting; IP: immunoprecipitation. (D) SUN5 interacted with CCDC113. pECMV-FLAG-Ccdc113 and pEGFP-GFP-Sun5 were transfected into HEK293T cells. At 48 h after transfection, the cells were collected for immunoprecipitation (IP) with anti-GFP antibody and analyzed with anti-FLAG and anti-GFP antibodies. (E) CENTLEIN interacted with CCDC113. pCDNA-FLAG-Centlein and pEGFP-GFP-Ccdc113 were transfected into HEK293T cells. At 48 h after transfection, the cells were collected for immunoprecipitation (IP) with anti-FLAG antibody and analyzed with anti-FLAG and anti-GFP antibodies. The % Input is displayed below the corresponding figures for quantification. n = 3 independent experiments. Data are presented as mean ± SD; *P < 0.05, ns indicates no significant difference. (G) Immunofluorescence of CCDC113 (red) and SUN5 (green) in mature spermatozoa. Nuclei were stained with DAPI (blue). (H) Immunofluorescence of CCDC113 (red) and CENTLEIN (green) in testicular step13-step14 spermatid. Nuclei were stained with DAPI (blue). (I) Immunofluorescence analysis for SPATA6 (green) and α/β-tubulin (red) was performed in Ccdc113+/+and Ccdc113−/− spermatozoa. Nuclei were stained with DAPI (blue). (J) Quantification ratio of SPATA6 on the detached sperm tail (n =3 independent experiments). At least 200 spermatozoa were analyzed for each mouse. (K) Quantification ratio of CCDC113 on the detached sperm tail (n = 3 independent experiments). At least 200 spermatozoa were analyzed from each mouse. Data are presented as mean ± SD; ***P < 0.001, ****P < 0.0001. (L) Immunofluorescence analysis for CCDC113 (red) was performed in wild type (WT), Sun5−/−, Centlein−/−, and Pmfbp1−/− spermatozoa. Nuclei were stained with DAPI (blue).

The HTCA localization of CCDC113 may be responsible for maintaining HTCA integrity. To test this hypothesis, we examined the localization of classical HTCA component SPATA6 in Ccdc113−/− and Ccdc113+/+ spermatozoa using immunofluorescent staining. We found that SPATA6 was not attached to the implantation fossa of the sperm nucleus in Ccdc113−/− spermatozoa (Figure 7I and J), indicating that CCDC113 is essential for the integrity of the sperm HTCA. To further elucidate the functional relationships among CCDC113, SUN5, CENTLEIN, and PMFBP1 at the sperm HTCA, we examined the localization of CCDC113 in Sun5-/-, Centlein−/−, and Pmfbp1−/− spermatozoa. Compared to the control group, CCDC113 was predominantly localized on the decapitated flagellum in Sun5-/-, Centlein−/−, and Pmfnp1−/− spermatozoa (Figure 7K and L), indicating SUN5, CENTLEIN, and PMFBP1 are crucial for the proper docking of CCDC113 to the implantation site on the sperm head. Taken together, these data demonstrate that CCDC113 cooperates with SUN5 and CENTLEIN to stabilize the sperm HTCA and anchor the sperm head to the tail.

Discussion

CCDC113 is a highly evolutionarily conserved component of motile cilia/flagella. Studies in the model organism, Tetrahymena thermophila, have revealed that CCDC113 connects RS3 to dynein g and the N-DRC, which plays essential role in cilia motility (Bazan et al., 2021; Ghanaeian et al., 2023). Recent studies have also identified the localization of CCDC113 within the 96-nm repeat structure of the human respiratory epithelial axoneme, and localizes to the linker region among RS, N-DRC and DMTs (Walton et al., 2023). In this study, we reveal that CCDC113 is indispensable for male fertility, as Ccdc113 knockout mice produce spermatozoa with flagellar defects and head-tail linkage detachment (Figure 3D). CCDC113 is localized to the sperm neck and flagellum regions in the elongating and elongated spermatids. In the sperm flagellum, CCDC113 interacts with both CFAP57 and CFAP91, serving as an adaptor protein to connect RS, N-DRC, and DMTs, thereby stabilizing sperm flagellum. At the sperm head to tail connecting piece, CCDC113 binds to SUN5 and CENTLEIN to stabilize sperm HTCA and anchor the sperm head to the tail. Thus, CCDC113 is essential for maintaining the integrity of both the sperm axoneme and the sperm HTCA. Moreover, TEM analysis detected excess residual cytoplasm in spermatozoa, including disrupted mitochondria, damaged axonemes, and large vacuoles, indicating defects in cytoplasmic removal in Ccdc113−/− mice (Figure 5A).

Recent cryo-EM analysis of the axonemes from Tetrahymena thermophila cilia and human respiratory cilia revealed that CCDC113 is localized at the linker region between RS3, N-DRC and DMTs (Bazan et al., 2021; Ghanaeian et al., 2023). We also found that CCDC113 interacts with adjacent axoneme-associated proteins, CFAP57 and CFAP91 (Figure 1C and D). CFAP57 has been identified as the adaptor protein responsible for assembling dynein g and d, and it can interact with both N-DRC and RS3 (Ghanaeian et al., 2023; Ma et al., 2023). Previous studies have demonstrated that CFAP91 extends from the base of RS2 through the N-DRC base plate to RS3, playing a crucial role in stabilizing and localizing RS2 and RS3 on the DMT (Bicka et al., 2022; Dymek et al., 2011; Gui et al., 2021). The CFAP91 ortholog, FAP91, interacts with three N-DRC subunits (DRC1, DRC2, and DRC4), facilitating the docking of the N-DRC in Chlamydomonas (Gui et al., 2021). In humans, pathogenic mutations in CFAP91 and DRC2 disrupt sperm flagellum structure and result in MMAF (Jreijiri et al., 2023; Martinez et al., 2020). TEM and immunostaining experiments in spermatozoa showed severe CP and radial spokes defects in CFAP91 mutant patients (Martinez et al., 2020). In this study, we found that the absence of CCDC113 results in severe axonemal disorganization characterized by defective RSs, scattered DMTs, and misplaced CP. Further analysis demonstrated that the CCDC113 deficiency disrupts the localization of CFAP91 and DRC2 on DMTs. Thus, CCDC113 may function as an adaptor protein to stabilize CFAP91 and DRC2 on DMTs, facilitating the docking of RS and N-DRC to DMTs and thereby maintaining the integrity of sperm axoneme.

Recent analyses have revealed that certain centrosomal proteins play crucial roles in the assembly and maintenance of sperm HTCA (Avasthi et al., 2013; Liska et al., 2009; Zhang et al., 2021). In early round spermatids, the centriole pair initially localizes to the caudal nuclear pole and expands the electron-dense material, some of which exhibits striation (Wu et al., 2020). As spermatids develop, the dense material around the centrioles gradually transforms into a well-organized structure, clearly identified as the basal plate, capitulum plate and segmented columns (Dooher & Bennett, 1973; Wu et al., 2020). CCDC113 was initially identified through the isolation of centriolar proteins from bovine sperm (Firat-Karalar et al., 2014). During spermiogenesis, CCDC113 colocalizes with CENTRIN1/2 at the centrosome in developing spermatids and continues to localize to the sperm neck region in elongating and elongated spermatids (Figure 4B and C). Additionally, CCDC113 can bind to HTCA associated centrosomal protein, CENTLEIN, and the disruption of Centlein impairs the attachment of CCDC113 to the sperm head. In Ccdc113−/−spermatids, the capitulum plate and segmented columns are absent, and the basal plate is detached from the implantation site on the nucleus of Ccdc113−/− elongating and elongated spermatids during spermiogenesis (Figure 6E). These observations raise the possibility that CCDC113 is an HTCA associated centrosomal protein crucial for maintaining the structural integrity of the HTCA.

SUN5 is a transmembrane protein located in the nuclear envelope, and acts as the root connecting the HTCA to the sperm nuclear envelope (Shang et al., 2017). CENTLEIN can directly bind to both SUN5 and PMFBP1, serving as a linker between SUN5 and PMFBP1 to maintain the integrity of HTCA (Zhang et al., 2021). CCDC113 can interact with SUN5 and CENTLEIN, but not PMFBP1 (Figure 7A-C), and left on the tip of the decapitated tail in Sun5−/− and Centlein−/− spermatozoa (Figure 7K and L). Furthermore, CCDC113 colocalizes with SUN5 in the HTCA region, and immunofluorescence staining in spermatozoa shows that SUN5 is positioned closer to the sperm nucleus than CCDC113 (Figure 7G and H). Therefore, SUN5 and CENTLEIN may be closer to the sperm nucleus than CCDC113. PAS staining revealed that Ccdc113−/− sperm heads are abnormally oriented in stages V–VIII seminiferous epithelia (Figure6 C and D), and TEM analysis further demonstrated that the disruption of CCDC113 causes the detachment of the destroyed coupling apparatus from the sperm head in step 9–11 spermatids (Figure 6E). All these results suggest that the detachment of sperm head and tail in Ccdc113−/− mice may not be a secondary effect of sperm flagellum defects.

Overall, we identified CCDC113 as a structural component of both the flagellar axoneme and the HTCA, where it performs dual roles in stabilizing the sperm axonemal structure and maintaining the structural integrity of HTCA. Given that the cryo-EM of sperm axoneme and HTCA could powerfully strengthen the role of CCDC113 in stabilizing sperm axoneme and head-tail coupling apparatus, it a valuable direction for future research. Ccdc113−/−mice did not exhibit other ciliopathies, such as situs inversus, hydrocephalus, or abnormal ciliogenesis of tracheal cilia (Figure 2-figure supplement 1B-E), which suggests that CCDC113 may specifically function in spermiogenesis.

Materials and methods

Phylogenetic analysis and structural similarity analysis

The amino acid sequence of CCDC113 of seven species were downloaded from UniProt. The phylogenetic trees were constructed using MEGA 10.0 (Kumar et al., 2018) with the Neighbor-Joining (NJ) method (Saitou & Nei, 1987). 3D structures of CCDC113 orthologs of seven species were obtained from AlphaFold Protein Structure Database (Varadi et al., 2022). Structural similarity Z scores were derived through the DALI webserver for all against all structure comparison (Holm & Laakso, 2016).

Animals

The Sun5−/−, Pmfbp1−/− and Centlein−/− mice have been reported previously (Shang et al., 2017; Zhang et al., 2021; Zhu et al., 2018). The mouse Ccdc113 gene is 44.81 kb and contains 9 exons. Exon 2 to exon 6 of Ccdc113 were chosen as the target sites. The knockout mice were generated using CRISPR-Cas9 system from Cyagen Biosciences. The gRNA and Cas9 mRNA were co-injected into fertilized eggs of C57BL/6J mice to generate Ccdc113+/−mice with a 10215 bp base deletion and 64 bp insertion. The resulting heterozygotes were interbred, and their offspring were genotyped by genomic DNA sequencing to identify wide-type and homozygote mice. The genotyping primers for knockout were: F1: 5’-TCAAATCATCACACCCTGCCTCT-3’, R: 5’-GCTTGCACTCGGGTGATACATAA-3’, and for WT mice, the specific primers were: F2: 5’-CAGGTTCTCAACACCTACAAAGTA-3’, R: 5’-GCTTGCACTCGGGTGATACATAA-3’. All the animal experiments were performed according to approved institutional animal care and use committee (IACUC) protocols (# 08-133) of the Institute of Zoology, Chinese Academy of Sciences.

Assessment of fertility

Fertility was assessed in 2-month-old male mice of various genotypes. Each male was paired with two WT females (6–8 weeks old), and vaginal plugs were checked each morning. Females with a vaginal plug were separated and housed individually, and their pregnancy outcomes were recorded. If a female did not give birth by day 22 postcoitus, it was considered not pregnant and euthanized for confirmation. Each male participated in six cycles of the breeding assay with different females.

Human adult sperm sample preparation

The sperm donation candidates in this study were healthy young Chinese men. Each participant underwent a thorough medical examination and completed an extensive medical and social questionnaire to ensure the exclusion of individuals with genetic or significant medical issues, as outlined in the Basic Standard and Technical Norms of Human Sperm Bank published by the Chinese Ministry of Health. Individuals who smoked, abused drugs, or were heavy drinkers were also excluded from the study. Those who remained eligible signed informed consent forms for voluntary sperm donation and agreed to reside in Beijing for a minimum of six months. The sperm bank documented each participant’s age, date of birth, and date of semen collection. Ethical approval for this study was granted by the Reproductive Study Ethics Committee of Peking University Third Hospital (2017SZ-035). Semen samples were processed using a 40% density gradient of PureSperm (Nidacon International, Molndal, Sweden) through centrifugation at room temperature (500 × g, 30 min) and washed three times with phosphate-buffered saline (PBS). The obtained spermatozoa were utilized for immunofluorescence staining.

Antibodies

Rabbit anti-CCDC113 generated from Dia-an Biotech (Wuhan, China) was diluted at 1:500 for western blotting and a 1: 25 dilution for immunofluorescence. Mouse anti-α-TUBULIN antibody (AC012, Abclonal) was used at a 1:5000 dilution for western blotting. Mouse anti-GFP antibody (M20004, Abmart) was used at a 1:2000 dilution for western blotting. Rabbit anti-MYC antibody (BE2011, EASYBIO) was used at a 1:2000 dilution for western blotting.

Rabbit anti-FLAG antibody (20543-1 AP, Proteintech) was used at a 1:2000 dilution for western blotting. Mouse anti-ac-TUBULIN antibody (T7451, Sigma-Aldrich) was used at a 1:200 dilution for immunofluorescence. Mouse anti-α/β-TUBULIN antibody (ab44928, Abcam) was used at a 1: 100 dilution for immunofluorescence. Rabbit anti-DRC2 antibody (NBP2-84617, Novus) was used at a 1:100 dilution for immunofluorescence. Rabbit anti-CFAP91 antibody (bs-9823R, Bioss) was used at a 1:100 dilution for immunofluorescence. In-house-generated mouse anti-SUN5 antibody targeting the SUN5 SUN domain (aa193–373) was used at a 1:100 dilution for immunofluorescence analysis. Rat anti-CENTLEIN antibody was generated by Absea Biotechnology Ltd (Beijing, China) was diluted at a 1:20 dilution for immunofluorescence. Rabbit anti-SPATA6 antibody (11849 −1 AP, Proteintech) was used at a 1:100 dilution for immunofluorescence. The Alexa Fluor 488 conjugate of lectin PNA (1:400, L21409, Thermo Fisher) was used for immunofluorescence. The secondary antibodies were goat anti-rabbit FITC (1:200, ZF-0311, Zhong Shan Jin Qiao), goat anti-rabbit TRITC (1:200, ZF-0316, Zhong Shan Jin Qiao), goat anti-mouse FITC (1:200, ZF-0312, Zhong Shan Jin Qiao) and goat anti-mouse TRITC (1:200, ZF0313, Zhong Shan Jin Qiao).

Sperm motility and sperm count assays

The cauda epididymis was isolated from the male mice of the different genotypes. Sperm were released in phosphate-buffered saline (PBS, Gibco, C14190500BT) from the incisions of the cauda epididymis for 10 min at 37°C. And then the swim-up suspension was used for the analysis of sperm motility with a microscope through a 20x phase objective. Viewing areas in each chamber were imaged using a CCD camera. The samples were analyzed via computer-assisted semen analysis (CASA) using the Minitube Sperm Vision Digital Semen Evaluation System (12500/1300, Minitube Group, Tiefenbach, Germany) and were also analyzed by CASA. The incubated sperm number was counted with a hemocytometer.

Histology staining

As previously reported (Wang et al., 2018), the testes and cauda epididymis were dissected after euthanasia, and fixed with Bouin’s fixative for 24 h at 4°C, then the testes were dehydrated with graded ethanol and embedded in paraffin. For histological analysis, the 5 μm sections were cut and covered on glass slides. Sections were stained with H&E and PAS for histological analysis after deparaffinization.

Electron microscopy analysis

The cauda epididymis and testis were dissected and fixed in 2.5% (vol/vol) glutaraldehyde in 0.1 M cacodylate buffer at 4°C overnight. After washing in 0.1 M cacodylate buffer, samples were cut into small pieces of approximately 1 mm3, then immersed in 1% OsO4 for 1 h at 4°C. Samples were dehydrated through a graded acetone series (50%, 60%, 70%, 80%, 90%, 95%, 100%) and embedded in resin (DDSA, NMA, enhancer, 812) for staining. Ultrathin sections were cut and stained with uranyl acetate and lead citrate. Images were acquired and analyzed using a JEM-1400 transmission electron microscope.

Scanning electron microscopy

The trachea was fixed in 2.5% glutaraldehyde solution overnight, and dehydrated in graded ethanol, subjected to drying and coated with gold. The images were acquired and analyzed using SU8010 scanning electron microscope.

Immunofluorescence staining

The testis albuginea was peeled and incubated with collagenase IV and hyaluronidase in PBS for 15 min at 37°C, then washed twice with PBS. Next, fixed with 4% PFA for 5 min, and then coated on slide glass to dry out. The slides were washed with PBS three times and then treated with 0.5% Triton X-100 for 5 min, and blocked with 5% BSA for 30 min. Added the primary antibodies and incubated at 4°C overnight, followed by incubating with a second antibody and DAPI. The images were taken using LSM880 and SP8 microscopes.

The mouse testis was immediately dissected and fixed with 2% paraformaldehyde in 0.05% PBST (PBS with 0.05% Triton X-100) at room temperature for 5 min. The fixed sample was placed on a slide glass and squashed by placing a cover slip on top and pressing down. The sample was immediately flash frozen in liquid nitrogen, and the slides were stored at −80°C for further immunofluorescence experiments. After removing the coverslips, the slides were washed with PBS three times and then treated with 0.1% Triton X-100 for 10 min, rinsed three times in PBS, and blocked with 5% bovine serum albumin (Amresco, AP0027). The primary antibody was added to the sections and incubated at 4°C overnight, followed by incubation with the secondary antibody. The nuclei were stained with DAPI. The immunofluorescence images were taken immediately using an LSM 780 microscope (Zeiss) or SP8 microscope (Leica).

Spermatozoa were released from the cauda epididymis in PBS at 37°C for 15 min, then were spread on glass slides for morphological observation or immunostaining. After air drying, spermatozoa were fixed in 4% PFA for 5 min at room temperature, and slides were washed with PBS three times and blocked with 5% BSA for 30 min at room temperature. The primary antibodies were added to the sections and incubated at 4°C overnight, followed by incubation with the secondary antibody. The nuclei were stained with DAPI and images were taken using an LSM 880 microscope (Zeiss) or SP8 microscope (Leica).

Immunoprecipitation

Transfected cells were lysed in a lysis buffer (50mM HEPES, PH 7.4, 250mM NaCl, 0.1% NP-40 containing PIC and PMSF) on ice for 30 min and centrifugated at 12000 rpm at 4°C for 15 min, cell lysates were incubated with primary antibody about 12 hours at 4°C and then incubated with protein A-Sepharose (GE, 17-1279-03) for 3 hours at 4°C, then washed three times with lysed buffer and subjected to immunoblotting analysis.

Statistical analysis

All the experiments were repeated at least three times, and the results are presented as the mean ± SD. The statistical significance of the differences between the mean values for the different genotypes was measured by the Student’s t-test with a paired, 2-tailed distribution. The data were considered significant when the P-value was less than 0.05(*), 0.01(**), 0.001(***) or 0.0001(****).